---
title: "Adding a New NTD Program"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a New NTD Program}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

::: {.eri-guide-meta}
**Walkthrough** &middot; ~20 min for the shipped example below &middot; needs: nothing &middot; sandbox-safe: [yes]{.eri-sandbox-yes} (local schema files, dry-run only)
:::

This vignette walks through the full process of adding a new disease program
to `erifunctions` -- from generating skeleton schemas to opening a pull request.
Estimated time for a *real* new country/disease combination (not just running the
example below): 1--2 hours.

## 1. Generate skeleton schemas

`eri_onboard_disease()` writes one YAML file per data type:

```{r scaffold}
library(erifunctions)

eri_onboard_disease(
  disease    = "schisto",
  country    = "ug",
  data_types = c("mda", "prevalence"),
  output_dir = "schemas/"     # write locally, not to the package
)
# Writes schemas/ug_schisto_programmatic_treatment.yaml and
#        schemas/ug_schisto_research_prevalence.yaml (ADR-0012: mda -> programmatic/
#        treatment, prevalence -> research/prevalence)
```

Use `dry_run = TRUE` first to preview what will be written:

```{r dry-run}
eri_onboard_disease("rb", "eth", dry_run = TRUE)
```

## 2. Fill in disease-specific columns

Open each skeleton file and complete the TODO sections:

```yaml
# ug_schisto_programmatic_treatment.yaml (excerpt)
columns:
  species_target:
    required: true
    type: categorical
    aliases: [Species, species, schisto_species]
    allowed_values:
      - S.mansoni        # fill in the species relevant to this country
      - S.haematobium

  school_name:           # add country-specific columns here
    required: false
    type: character
    aliases: [School, school_name, etablissement]
```

Key decisions to make for each column:

- **`required`** -- will the pipeline hard-error if this column is missing?
- **`type`** -- `numeric`, `character`, `categorical`, or `date`
- **`aliases`** -- all the column name variants found in raw data files
- **`range`** -- for numeric columns; flag implausible values
- **`allowed_values`** -- for categorical columns; triggers a flag if a value is unexpected
- **`translations`** -- map raw variants to a canonical value (e.g. `"YES"` -> `"Yes"`)

## 3. Validate the schema

Run `eri_schema_validate()` to check for structural problems:

```{r validate}
result <- eri_schema_validate("schemas/ug_schisto_mda.yaml")
# Prints: "Schema ug_schisto_mda.yaml is valid." if no issues found
# Or:     a summary of issues with field names and messages
print(result)
```

Fix any flagged issues before moving on. Common problems:

- Missing `required` or `type` on a column definition
- A consistency rule referencing a column name not defined in `columns`
- A `temporal.year_col` or `temporal.period_col` pointing to a column not in `columns`

## 4. Test with sample data

Before submitting, test the schema against real or synthetic data:

```{r test-dq}
schema <- load_dq_schema("ug", "schisto", "programmatic", "treatment", azcontainer = NULL)
# azcontainer = NULL tells it to use the local file, not Azure

# Create a small synthetic data frame that matches your schema
sample_data <- data.frame(
  year       = c(2023L, 2023L, 2024L),
  round      = c(1L, 2L, 1L),
  district   = c("Kampala", "Wakiso", "Kampala"),
  target_pop = c(5000L, 3200L, 5100L),
  treated    = c(4800L, 3100L, 4950L),
  coverage_pct = c(96.0, 96.9, 97.1)
)

result <- run_dq_checks(sample_data, schema)
cat("Flags:", nrow(result$flags), "\n")
```

If the sample data passes cleanly (`nrow(result$flags) == 0`), the schema is
ready for real data.

## 5. Submit to the package

Copy the validated schema file to `inst/schemas/` using the four-part identity
`{country}_{disease}_{data_source}_{data_type}.yaml` (ADR-0012):

```
inst/schemas/ug_schisto_programmatic_treatment.yaml
inst/schemas/ug_schisto_research_prevalence.yaml
```

PR checklist:

- [ ] Schema file is in `inst/schemas/` with correct name
- [ ] `eri_schema_validate()` returns 0 issues
- [ ] `load_dq_schema("ug", "schisto", "programmatic", "treatment", azcontainer = NULL)` works
- [ ] At least one `load_dq_schema` + `eri_schema_validate` test added to
  `tests/testthat/test-onboarding.R` following the existing pattern
- [ ] NEWS.md entry in the next version section

## 6. Optional: add disease-specific epi functions

If the new program needs custom analytics beyond the generic DQ pipeline
(e.g. DALY calculations, TAS interpretation, programme status maps), add
them to a new file `R/epi_{disease}.R` following the pattern in
`R/epi_lf.R`:

```{r epi-template}
# R/epi_schisto.R

#' Compute school-level Kato-Katz prevalence from survey data
#'
#' @param data A tibble with one row per tested individual.
#' @param result_col `chr` Column containing `"Positive"` / `"Negative"`.
#' @param group_cols `chr` vector Grouping columns (school, district, year).
#' @returns A tibble of prevalence estimates by group.
#' @export
eri_schisto_kk_prevalence <- function(data, result_col, group_cols) {
  # implementation here
}
```

Export and document following standard roxygen conventions, then add tests in
`tests/testthat/test-epi-schisto.R`.
