---
title: "Connecting to Azure, ODK Central, SharePoint, and Teams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting to Azure, ODK Central, SharePoint, and Teams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`erifunctions` talks to four outside services on your behalf — **Azure** (where the data lives),
**ODK Central** (where field submissions come from), **SharePoint** (shared files), and **Teams**
(notifications). This is the one place that shows how to connect to each and **confirm it works**.

The good news: most need little or no setup. Azure and SharePoint are **zero-config** — you just sign
in through your browser. ODK needs three lines in a settings file. Teams is optional. Work through the
service(s) you need; skip the rest.

> **A note on secrets.** Where a service needs a credential, it lives in your **`.Renviron`** file —
> never typed into a script and never committed to git. We point that out at each step.

```{=html}
<div class="mermaid">
flowchart TD
    You["Your R session"] --> P["erifunctions"]
    P --> AZ["Azure Storage — where the data lives"]
    P --> ODK["ODK Central — field submissions"]
    P --> SP["SharePoint — shared files"]
    P --> TM["Teams — notifications"]
</div>
```

## At a glance

| Service | Connect with | How you sign in | Setup needed | Confirm it works |
|---|---|---|---|---|
| **Azure** | `get_azure_storage_connection()` | browser (your Carter Center account) | **none** | `eri_list()` |
| **ODK Central** | `init_odk_connection()` | email + password | 3 lines in `.Renviron` | `list_odk_projects()` |
| **SharePoint** | `eri_sharepoint_connect(site_url)` | browser | site URL only | `eri_sharepoint_list()` |
| **Teams** | `get_teams_connection()` | token / device code / webhook | a token *or* a webhook | `eri_teams_send(message = "…")` |

```{r library}
library(erifunctions)
```

## Azure — nothing to configure {#azure}

This is the big one: **the data system itself**. And there is nothing to set up. Thanks to baked-in
defaults (see `?get_azure_storage_connection` and ADR-0008), the **first time** a command needs Azure
your browser opens, you sign in with your Carter Center account, and the sign-in is remembered for the
rest of the session.

```{r azure-connect}
# The first call this session opens your browser to sign in.
# Nothing prints on success — you get a connection object back.
data_con <- get_azure_storage_connection(storage_name = "data")
```

The `storage_name` picks **which blob you connect to**: `"data"` (the governed three-layer data
system) or `"projects"` (the legacy contractor space). Most analyst work is in `"data"`.

**Confirm it works** by listing the top of the container — if you get a tibble back, you are in:

```{r azure-verify}
eri_list("", azcontainer = data_con)
#> # A tibble: 5 × 4
#>   name      size isdir lastModified
#>   <chr>    <dbl> <lgl> <dttm>
#> 1 dr          NA TRUE  2026-01-15 09:12:04
#> 2 ht          NA TRUE  2026-02-01 14:30:51
#> 3 uga         NA TRUE  2026-03-22 11:05:18
#> …
```

> **For automation / CI (headless).** A scheduled job has no browser. Set a **service principal** —
> `ERIFUNCTIONS_SP_CLIENT_ID` and `ERIFUNCTIONS_SP_CLIENT_SECRET` in the environment (or pass
> `creds_yaml_path=`) — and `get_azure_storage_connection()` signs in non-interactively instead. The
> secret is the *only* true secret here; the tenant, app, and endpoint are non-sensitive defaults.

> **`403 Forbidden`?** That is not a bug — it is Azure telling you your account does not have access to
> that storage. Access is granted by an administrator (Azure RBAC), not by the package. Ask your ERI
> admin to add you.

## ODK Central — three lines in your settings file {#odk}

ODK Central signs in with your **email and password**, so those go in your `.Renviron` once. Open it
with:

```{r odk-renviron}
usethis::edit_r_environ()
```

Add three lines, then **save and restart R**:

```
ODK_URL=https://your-odk-server.org/
ODK_USER=you@example.org
ODK_PASS=your-password
```

Now connect — `init_odk_connection()` reads those automatically, so your password never appears in a
script:

```{r odk-connect}
con <- init_odk_connection()
#> ✔ Connected to <https://your-odk-server.org/>. Session expires 2026-06-26T20:00:45.690Z.
```

**Confirm it works** by listing the projects you can see:

```{r odk-verify}
list_odk_projects(con = con)
#> # A tibble: 3 × 3
#>   project_id project   description
#>        <int> <chr>     <chr>
#> 1          5 Uganda    NA
#> 2          7 Training  NA
#> 3         11 testing   NA
```

Pass that `con` to the other ODK functions as `con = con`. For the full ODK workflow — registering a
form, monitoring it, and pulling submissions — see the
[ODK Central guide](da-odk-guide.html).

> **For automation / CI.** Instead of email+password, you can reuse a **bearer token**: set `ODK_TOKEN`
> in the environment and call the ODK functions with `con = NULL` — they fall back to `ODK_URL` +
> `ODK_TOKEN`.

## SharePoint — just the site URL {#sharepoint}

SharePoint uses the same browser sign-in as Azure (via `Microsoft365R`), so all you supply is the
**site URL**:

```{r sharepoint-connect}
site <- eri_sharepoint_connect("https://cartercenter.sharepoint.com/sites/ERI")
# Browser opens on first use; the token is cached for the session.
```

**Confirm it works** by listing a folder in the site's document library:

```{r sharepoint-verify}
eri_sharepoint_list(site, "Shared Documents/Malaria/2024")
#> # A tibble: 4 × 5
#>   name              size modified            is_folder path
#>   <chr>            <dbl> <dttm>              <lgl>     <chr>
#> 1 summary.xlsx     20481 2026-04-02 08:15:00 FALSE     Shared Documents/Malaria/2024/summary.xlsx
#> 2 maps             NA    2026-03-28 16:40:00 TRUE      Shared Documents/Malaria/2024/maps
#> …
```

For reading and uploading files, see `eri_sharepoint_read()` / `eri_sharepoint_upload()` and the
[SharePoint workflow](sharepoint-workflow.html) vignette.

> **No service-principal path.** SharePoint is browser-sign-in only in `erifunctions` — there is no
> headless/CI path for it. Use it interactively.

## Teams — for posting notifications {#teams}

Teams is **optional** — you only need it if you want `erifunctions` to post messages (for example,
[`eri_notify_dq()`](../reference/eri_notify_dq.html) posts a data-quality summary to a channel). There
are three ways in, easiest first:

**The simplest: an incoming webhook.** Ask a channel owner for an *incoming webhook URL*, put it in
your `.Renviron`, and you can post to that channel with no token at all:

```
ERIFUNCTIONS_TEAMS_WEBHOOK=https://outlook.office.com/webhook/…
```

```{r teams-webhook}
eri_teams_send(message = "Hello from erifunctions.")
#> ℹ Sending Teams message via incoming webhook.
```

**For direct messages, use the Graph API.** `get_teams_connection()` returns a token — from
`ERIFUNCTIONS_TEAMS_TOKEN` if you have one, otherwise via an interactive device-code sign-in. Unlike
Azure, this path is **not** zero-config: the device-code flow needs `ERIFUNCTIONS_APP_ID` set in your
`.Renviron` (without it you'll get a "No Teams token or app ID found" warning). Most people only need
to post to a channel, which the webhook above already covers.

```{r teams-token}
token <- get_teams_connection()
```

**Confirm it works** by sending yourself a DM:

```{r teams-verify}
eri_teams_send(message = "Test message to myself.", to = "self", token = token)
#> ℹ Sending Teams DM via Graph API.
```

> **For automation / CI.** A webhook (`ERIFUNCTIONS_TEAMS_WEBHOOK`) is the most robust for scheduled
> jobs — it needs no interactive sign-in. Where conditional-access policies block the device-code flow,
> the webhook is the way through. For DMs in automation, set a pre-obtained `ERIFUNCTIONS_TEAMS_TOKEN`.

## One `.Renviron` for everything {#renviron}

All your settings live in one file. Open it with `usethis::edit_r_environ()` and keep only the lines
you actually need:

```
# --- Azure storage --------------------------------------------------------
# Optional for analysts: tenant / app / endpoint have working built-in defaults.
# ERIFUNCTIONS_TENANT_ID=<Azure tenant ID>
# ERIFUNCTIONS_APP_ID=<Azure app registration ID>
# ERIFUNCTIONS_RESOURCE_ENDPOINT=<storage account endpoint URL>
ERIFUNCTIONS_STORAGE_NAME=projects
ERIFUNCTIONS_DATA_STORAGE_NAME=data

# Your identity (appears in approval and access logs)
ERI_ANALYST_ID=firstname.lastname

# --- ODK Central ----------------------------------------------------------
ODK_URL=https://your-odk-server.org/
ODK_USER=you@example.org
ODK_PASS=your-password

# --- Teams (optional) -----------------------------------------------------
# ERIFUNCTIONS_TEAMS_WEBHOOK=https://outlook.office.com/webhook/…

# --- Automation / CI only -------------------------------------------------
# ERIFUNCTIONS_SP_CLIENT_ID=<service principal client ID>
# ERIFUNCTIONS_SP_CLIENT_SECRET=<service principal client secret>
```

**Always restart R after editing `.Renviron`.** Everything in this file is read once at startup — and
none of it is ever written into your scripts or committed to git.

## Troubleshooting {#troubleshooting}

| Symptom | What it means | Fix |
|---|---|---|
| Azure `403 Forbidden` | Your account lacks access to that storage (RBAC) | Ask an ERI admin to grant access — it is not a code problem |
| ODK: *"ODK username is required"* | `ODK_USER` / `ODK_PASS` aren't set | Add them to `.Renviron`; restart R |
| A call re-opens the browser unexpectedly | Your cached session expired | Just sign in again — connections are session-scoped |
| Teams device-code sign-in is blocked | Conditional-access policy | Use an incoming webhook (`ERIFUNCTIONS_TEAMS_WEBHOOK`) instead |
| SharePoint "package Microsoft365R is required" | Optional dependency missing | `install.packages("Microsoft365R")` |

## What's next

Now that you can connect, the role guides put these connections to work:

- [Ingesting a surveillance dataset](da-ingest-guide.html) — the `raw → staged → approved` pipeline.
- [Working with ODK Central](da-odk-guide.html) — register, monitor, and pull form data.
- [A complete research workflow for epidemiologists](epi-research-guide.html) — running a study.

For every connection function grouped together, see the **Connections & authentication** section of
the [reference](../reference/index.html).
