---
title: "Answering ad-hoc data requests with SQL (eri_query)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Answering ad-hoc data requests with SQL (eri_query)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

"How many malaria cases by province this year?" "Pull treatment coverage for these three countries."
Ad-hoc requests are a daily part of the job, and they usually mean the same thing: **combine a few
approved datasets and compute a number.** [`eri_query()`](../reference/eri_query.html) does that with
**SQL**, so you don't hand-read and merge parquet files one at a time.

## The golden rule

> **`eri_query()` runs SQL across *processed* data — the approved, trustworthy layer — without moving it
> out of Azure or standing up a database.** It attaches the relevant parquet into an in-process
> [DuckDB](https://duckdb.org) session, runs your query, and hands back a tibble.

```{=html}
<div class="mermaid">
flowchart LR
    C["Approved data (processed/)"] --> Q["eri_query(sql, ...)"]
    Cat["Data catalog (which files to attach)"] --> Q
    Q --> R["a tibble"]
    R --> O["table / figure / answer"]
</div>
```

## Before you start {#before-you-start}

```{r install}
# eri_query needs two optional packages, installed once:
install.packages(c("duckdb", "DBI"))
library(erifunctions)
```

Querying approved data in Azure also needs a connection (the browser opens on first use); pure
in-memory queries — the examples in [Joins and aggregations](#explicit) below — need nothing.

```{r connect}
data_con <- get_azure_storage_connection(storage_name = "data")
```

## Two ways to put data in scope

`eri_query()` gives you two ways to choose what your SQL can see, and they combine:

1. **Catalog-driven** — name the data by its axes (`country`, `disease`, `data_type`, …) and
   `eri_query()` finds the matching **processed** files, stamps each row with its
   `country`/`disease`/`data_source`/`data_type`/`period`, and unions them into one table called
   `data`. This is how you roll up across countries or periods.
2. **Explicit tables** — hand it named tables directly (`tables = list(name = …)`), where each is a
   data.frame, a local `.parquet`, or a `data/` blob path. This is how you join your own inputs (a
   population table, a lookup) to approved data.

## Roll up across the catalog {#catalog}

The catalog is the index of approved data; `eri_query()` reads it for you. To total malaria cases by
country from every approved aggregate dataset:

```{r catalog-rollup}
eri_query(
  "SELECT country, SUM(total_cases) AS cases
     FROM data GROUP BY country ORDER BY cases DESC",
  disease = "malaria", data_type = "aggregate", data_con = data_con
)
#> ℹ Querying 1 table: "data".
#> ✔ Returned 2 rows.
#> # A tibble: 2 × 2
#>   country  cases
#>   <chr>    <dbl>
#> 1 ht      12345
#> 2 dr       9876
```

Because every row is stamped with its provenance, you can `GROUP BY country`, `period`, `data_type`,
etc. without joining anything. (Those five names are reserved in this mode — see
`?eri_query`.) Not sure what's available? [`eri_catalog_query()`](../reference/eri_catalog_query.html)
with the same filters lists the files that would be attached.

## Joins and aggregations {#explicit}

The explicit mode runs entirely in memory, so these you can run right now. Say a request lands as two
small tables — weekly case counts and a population denominator:

```{r data}
cases <- data.frame(
  province = c("North", "North", "South", "South", "East"),
  epiweek  = c(1, 2, 1, 2, 1),
  cases    = c(40, 55, 12, 9, 30)
)
pop <- data.frame(
  province = c("North", "South", "East"),
  pop      = c(120000, 45000, 80000)
)
```

**Aggregate** — total cases by province:

```{r agg}
eri_query(
  "SELECT province, SUM(cases) AS cases
     FROM cases GROUP BY province ORDER BY cases DESC",
  tables = list(cases = cases)
)
#> ℹ Querying 1 table: "cases".
#> ✔ Returned 3 rows.
#> # A tibble: 3 × 2
#>   province cases
#>   <chr>    <dbl>
#> 1 North       95
#> 2 East        30
#> 3 South       21
```

**Join** — incidence per 1,000, combining the two tables:

```{r join}
eri_query(
  "SELECT c.province, SUM(c.cases) * 1000.0 / p.pop AS incidence_per_1000
     FROM cases c JOIN pop p USING (province)
     GROUP BY c.province, p.pop ORDER BY incidence_per_1000 DESC",
  tables = list(cases = cases, pop = pop)
)
#> ℹ Querying 2 tables: "cases" and "pop".
#> ✔ Returned 3 rows.
#> # A tibble: 3 × 2
#>   province incidence_per_1000
#>   <chr>                 <dbl>
#> 1 North                 0.792
#> 2 South                 0.467
#> 3 East                  0.375
```

**Window functions** — week-over-week change, which is awkward in base R but one line of SQL:

```{r window}
eri_query(
  "SELECT province, epiweek, cases,
          cases - LAG(cases) OVER (PARTITION BY province ORDER BY epiweek) AS wow_change
     FROM cases ORDER BY province, epiweek",
  tables = list(cases = cases)
)
#> ℹ Querying 1 table: "cases".
#> ✔ Returned 5 rows.
#> # A tibble: 5 × 4
#>   province epiweek cases wow_change
#>   <chr>      <dbl> <dbl>      <dbl>
#> 1 East           1    30         NA
#> 2 North          1    40         NA
#> 3 North          2    55         15
#> 4 South          1    12         NA
#> 5 South          2     9         -3
```

You can mix the modes — pass catalog filters **and** `tables = list(pop = …)` to join approved cases to
a population table you supply.

## Turn the answer into an output

`eri_query()` returns an ordinary tibble, so the result flows straight into the rest of the toolkit —
[`eri_table()`](../reference/eri_table.html) for a branded table, `ggplot2` / the `eri_map_*` helpers
for a figure, or `eri_case_summary()` / `eri_incidence_rate()` for standard epi summaries. The
[reporting helpers](da-ingest-guide.html) take it from there.

## Tips

- **It queries *approved* data.** `eri_query()` reads the `processed/` layer through the catalog — the
  data the team trusts — not `raw/` or `staged/`.
- **Big scans are bounded by memory/download** (ADR-0004): filter tight (`country`/`disease`/`period`)
  so you attach only what you need.
- **It's DuckDB SQL** — joins, `GROUP BY`, window functions, CTEs, date functions all work; see the
  [DuckDB docs](https://duckdb.org/docs/sql/introduction).

## What's next

- The [data catalog reference](../reference/eri_catalog_query.html) — see what approved data exists
  before you query it.
- The [surveillance ingest guide](da-ingest-guide.html) — how data *gets* approved in the first place.
- The [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md).
```
