---
title: "Uploading and processing a monthly country report (CMR) (for data analysts)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uploading and processing a monthly country report (CMR) (for data analysts)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Every month, country programmes file a **Case Management Report (CMR)**, a filled Excel template of
treatment, training, and survey numbers. This guide, for a **Data Analyst**, walks the monthly job:
take that incoming Excel, **upload** it to Azure, **stage** it, **parse** each sheet, and **approve** it
into the governed data system.

The trick that makes it manageable: every CMR template has a row of **machine-readable field codes**
(`#rbtrt_year`, `#rbtrt_treated`, …). Those codes are identical across countries *and languages*, so
one function parses an English Ugandan template and a French Chadian one exactly the same way.

> **The golden rule is the same as every other dataset.** The report moves
> `projects/ (as filed)` → `staged/` → `processed/`, and nothing becomes canonical until you
> [`eri_approve()`](../reference/eri_approve.html) it. This is the CMR sibling of the
> [surveillance ingest guide](da-ingest-guide.html).

```{=html}
<div class="mermaid">
flowchart TD
    A["Country files the monthly CMR Excel"] --> B["Upload to the projects blob"]
    B --> C["eri_stage_cmr(): projects -> staged/"]
    C --> D["eri_ingest_cmr(): parse each sheet by its #field codes"]
    D --> E["Review the numbers"]
    E --> F["eri_split_cmr(): route each sheet to {disease}/programmatic/{measure}/staged/"]
    F --> G["eri_approve() each disease/measure -> processed/ + catalog"]
</div>
```

## Before you start {#before-you-start}

- `remotes::install_github("thecartercenter/erifunctions")`.
- **Azure access** for the upload / stage / approve steps (zero-config browser sign-in, see the
  [connections guide](connections-guide.html)). The **parsing** step (`eri_ingest_cmr()`) is offline,
  it just reads a local Excel file.

```{r library}
library(erifunctions)
```

> **A note on the examples below.** The upload, stage, and approve steps run against the live Azure
> `projects`/`data` blobs and real, registered countries, so the outputs for those are shown as
> illustrations rather than run here. Unlike the surveillance and ODK guides, there is **no
> throwaway-sandbox path** for stage/approve: `eri_stage_cmr()` only accepts a registered RB-expansion
> country (it rejects a made-up one). The **parsing** step *is* run for real on a small **synthetic**
> example report that ships with the package, so that output is exactly what you'll see.

## 1. Where the report comes in {#arrives}

A filled monthly template lands in the **`projects` blob**, under a per-country, per-period folder.
The period is a six-digit `YYYYMM`:

```
projects: health-rb-country-expansion-dev/raw/filled_templates/{country}/{period}/{file}.xlsx
e.g.      …/raw/filled_templates/uga/202406/uga_cmr_2024_06.xlsx
```

Do this from R with `eri_upload()`, it keeps the whole monthly run in one place and one tool. (If a
file already arrived via Azure Storage Explorer or SharePoint, that's fine too; you can skip ahead to
staging.) Connect to the `projects` blob and upload the file to that path:

```{r upload}
# Connect to the projects blob explicitly (the same blob eri_stage_cmr reads from).
projects_con <- get_azure_storage_connection(storage_name = "projects")

eri_upload(
  "uga_cmr_2024_06.xlsx",
  "health-rb-country-expansion-dev/raw/filled_templates/uga/202406/uga_cmr_2024_06.xlsx",
  azcontainer = projects_con
)
# (uploads the file; no console output)
```

### What the template looks like

Every CMR sheet has the same shape, a few rows of human-readable headers, then the **field-code row**,
then the data:

| Excel row | Contents |
|---|---|
| 1–4 | Title, group headers, human-readable column names |
| **5** | **`#field codes`** (e.g. `#rbtrt_year`, `#rbtrt_adm1`, `#rbtrt_treated`), the parsing anchor |
| 6+ | The monthly numbers |

That row 5 is what makes the whole thing work, as you'll see in §3.

## 2. Stage it {#stage}

`eri_stage_cmr()` pulls the filed Excel(s) out of the `projects` blob and copies them into the
governed `data` blob's `staged/` layer, ready for review. Give it the country and the period:

```{r stage}
eri_stage_cmr("uga", "202406")
#> ✔ Staged: uga_cmr_2024_06.xlsx
#> ── ✔ Staged CMR to data blob ──────────────────────────────
#> Files: 1
#> Location: uga/rblf/cmr/staged
#> ℹ Operation log: uga/rblf/cmr/logs/20240705_141500_eri_stage_cmr_202406.yaml
```

Omit the period and it stages the **most recent** one for you, and says so:

```{r stage-auto}
eri_stage_cmr("uga")
#> ℹ No period specified; staging most recent: 202406
#> ✔ Staged: uga_cmr_2024_06.xlsx
#> …
```

(CMR data lives under the `rblf` disease folder, RB for onchocerciasis, LF for lymphatic filariasis,
the two programmes these reports cover. Only the registered RB-expansion countries,
`eth`, `nga`, `sdn`, `ssd`, `uga`, `mad`, `tcd`, can be staged.)

> **Heads-up on the `rblf` coordinate.** CMR always uses `disease = "rblf"`, the combined RB + LF
> programme code, so the paths are `{country}/rblf/cmr/…` and you approve with
> `eri_approve(country, "rblf", "cmr", period)`. Scaffolding a new country's CMR with
> `eri_onboard_cmr(create_dirs = TRUE)` creates those same `{country}/rblf/cmr/` folders.
>
> **From `rblf`/`cmr` to per-disease.** `rblf`/`cmr` were interim coordinates: under the
> [source ≠ measure model (ADR-0012)](https://github.com/thecartercenter/erifunctions/blob/main/docs/adr/0012-source-measure-data-model.md),
> CMR is a *format* of the `programmatic` channel, and [`eri_split_cmr()`](#split) (below) routes each
> sheet to its own disease and measure, e.g. RB Treatment → `{country}/oncho/programmatic/treatment/`.
> You then approve **each disease/measure**, not one combined `rblf`/`cmr` bucket.

## 3. Parse each sheet {#parse}

Now the offline heart of the job. First, see which sheets this country's template has:

```{r schema}
schema <- load_cmr_schema("uga")
names(schema$sheets)
#>  [1] "RB Treatment"                  "SCH Treatment"
#>  [3] "LF MMDP"                       "VHT Training"
#>  [5] "Parish Supervisors Training"   "Local Leaders Training"
#>  [7] "Subcounty Supervisor Training" "MMDP (surgery) Training"
#>  [9] "MMDP (patient) Training"       "Field Ento Training"
#> [11] "Lab Training"                  "LF Surveys"
#> [13] "RB Epi Surveys"                "RB Ento Surveys"
```

(A real monthly file has many sheets, treatments, MMDP, the various trainings, and surveys. The
bundled `cmr-example.xlsx` below is a small **two-sheet** subset, `RB Treatment` + `SCH Treatment`, so
you can run the parse and split offline.)

Then read a sheet with [`eri_ingest_cmr()`](../reference/eri_ingest_cmr.html). It reads from **row 5**
(the field codes), keeps only the `#`-coded columns, drops the template's blank spacer rows, and, when
you pass `country`, tags each row with it. (We use the package's bundled synthetic example here;
in real work you'd point at your staged file.)

```{r ingest-rb}
report <- system.file("extdata", "cmr-example.xlsx", package = "erifunctions")

rb <- eri_ingest_cmr(report, sheet = "RB Treatment", country = "uga")
#> ✔ CMR sheet "RB Treatment": 3 data rows, 6 field codes.

rb
#> # A tibble: 3 × 7
#>   country `#rbtrt_year` `#rbtrt_month` `#rbtrt_adm1` `#rbtrt_adm2` `#rbtrt_target` `#rbtrt_treated`
#>   <chr>   <chr>         <chr>          <chr>         <chr>         <chr>           <chr>
#> 1 uga     2024          06             Central       Kampala       12000           11400
#> 2 uga     2024          06             Western       Mbarara       8000            7600
#> 3 uga     2024          06             Eastern       Soroti        6000            5800
```

Each sheet is its own programme, parsed the same way:

```{r ingest-sch}
sch <- eri_ingest_cmr(report, sheet = "SCH Treatment", country = "uga")
#> ✔ CMR sheet "SCH Treatment": 2 data rows, 6 field codes.
```

The column names *are* the field codes, stable identifiers you can rely on across every monthly file.
This is the moment to eyeball the numbers (treated vs target, missing districts) before you sign off.

> **Same codes, any language.** The field codes (`#rbtrt_…`) are language-neutral, so the *same*
> `eri_ingest_cmr()` parses a French template too. For a French file, the sheet is named in French
> (e.g. `"Oncho Traitement"`), pass the **canonical slug** and the country, and the schema's
> `sheet_aliases` resolves it:
>
> ```r
> eri_ingest_cmr("tcd_cmr_2024_06.xlsx", sheet = "rb_treatment", country = "tcd")
> #> ✔ CMR sheet "Oncho Traitement": 24 data rows, 6 field codes.
> ```

## 4. Split it by disease and measure {#split}

The sheets are per-programme, but the canonical store is **per-disease**.
[`eri_split_cmr()`](../reference/eri_split_cmr.html) reads every routable sheet and routes its rows to
`{country}/{disease}/programmatic/{measure}/staged/`: the **disease from the sheet** (RB Treatment →
`oncho`, SCH Treatment → `sch`, LF MMDP → `lf`), the **measure from its category**. It reads the Excel
**directly** (here the bundled example; in real work the file you uploaded), independently of the
`rblf/cmr/staged/` copy `eri_stage_cmr()` made, that copy is the raw archive; this is the parse + route.
Preview the routing first with `dry_run`:

```{r split-dry}
eri_split_cmr(report, "uga", dry_run = TRUE)
#> ✔ CMR sheet "RB Treatment": 3 data rows, 6 field codes.
#> ✔ CMR sheet "SCH Treatment": 2 data rows, 6 field codes.
#> ℹ Dry run -- nothing written. Routing plan:
#>   "RB Treatment"  -> uga/oncho/programmatic/treatment/staged/cmr-example_rb_treatment.parquet (3 rows)
#>   "SCH Treatment" -> uga/sch/programmatic/treatment/staged/cmr-example_sch_treatment.parquet (2 rows)
#> Warning: Sheet "LF MMDP" not found in cmr-example.xlsx; skipping.   # ...and the
#> Warning: Sheet "VHT Training" not found in cmr-example.xlsx; skipping.  # other
#> # ... (the example has only the two treatment sheets; a real file routes them all)
```

A sheet the schema routes but the workbook doesn't contain is **skipped with a warning**, expected
here, since the example is a two-sheet subset. (If *none* of the routable sheets are present,
`eri_split_cmr()` errors instead of silently doing nothing.)

> **Why the sheet, not the row?** Each sheet carries a per-row `#…_disease` field whose values are
> *programme-coverage* codes (`RB` / `RBLF` / `RBLFSCH`, which programmes run at that location), **not**
> a single disease. `eri_split_cmr()` keeps that field as a data column and takes the routing disease
> from the sheet, so the same treatment numbers are **never duplicated** across diseases. Cross-programme
> **Training** sheets, which serve all programmes at once, route together under the combined `rblf` disease.

Then stage for real, one parquet per sheet, in its disease/measure folder:

```{r split-real}
eri_split_cmr(report, "uga")
#> ✔ "RB Treatment"  -> uga/oncho/programmatic/treatment/staged/cmr-example_rb_treatment.parquet
#> ✔ "SCH Treatment" -> uga/sch/programmatic/treatment/staged/cmr-example_sch_treatment.parquet
#> ── ✔ Split CMR by disease/measure ───────────────────────
#> Sheets: 2 routed
#> Diseases: oncho, sch
```

## 5. Approve each disease/measure {#approve}

Each split dataset passes through the same human gate, once per disease/measure:

```{r approve}
eri_approve("uga", "oncho", "programmatic", "202406", data_type = "treatment")
#> ✔ Approved: cmr-example_rb_treatment.parquet
#> ── ✔ Approved "202406" ──────────────────────────────────
#> Dataset: uga / oncho / programmatic / treatment
#> Location: uga/oncho/programmatic/treatment/processed

eri_approve("uga", "sch", "programmatic", "202406", data_type = "treatment")
#> ── ✔ Approved "202406" ──────────────────────────────────
#> Dataset: uga / sch / programmatic / treatment
```

Each disease's data is now canonical and discoverable in the catalog on its own coordinates:

```{r catalog}
eri_catalog_query(country = "uga", data_source = "programmatic")
#> # A tibble: 2 × 13
#>   path                              country disease data_source  data_type layer …
#>   <chr>                             <chr>   <chr>   <chr>        <chr>     <chr>
#> 1 uga/oncho/programmatic/treatmen…  uga     oncho   programmatic treatment proce…
#> 2 uga/sch/programmatic/treatment/…  uga     sch     programmatic treatment proce…
```

That's the monthly loop: **upload → stage → split → approve.**

## What's next

- **A new country files its first report?** Its CMR schema (which sheets, which field codes) has to
  exist first, see the CMR section of the [onboarding guide](da-onboard-guide.html)
  (`eri_onboard_cmr()`).
- **`eri_split_cmr()` says "No routable sheets"?** A sheet only routes when its schema entry declares a
  `disease` and a `data_type`. `uga` is the worked example; other countries' routing keys are being
  filled in, add them to the country's `inst/schemas/cmr/{code}.yaml` (one `disease` + `data_type` per
  sheet) to enable the split there.
- The [surveillance ingest guide](da-ingest-guide.html) covers the same `raw → staged → approved` gate
  for line-list data, and the data-catalog mechanics in more depth.

> **Real reports are not practice data.** The monthly CMR files are protected country data, staged and
> approved through this pipeline, never deleted or moved casually. (The example here is synthetic, which
> is the only reason we could pass it around freely.)

See the [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md) for the
full set of guides.
