---
title: "eri_do(): a tour of the guided console wizard (for data analysts)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eri_do(): a tour of the guided console wizard (for data analysts)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

::: {.eri-guide-meta}
**Desk reference** &middot; ~15 min &middot; needs: n/a to read, Azure/ODK to actually run &middot; sandbox-safe: [yes]{.eri-sandbox-yes} (illustrated on `atlantis`)
:::

`eri_do()` is the guided console wizard: run it, pick what you're doing from a menu, and answer a
few prompts, names you already know, files you already have, yes/no decisions, never a function
name or an Azure path. This is a **tour of the wizard itself**: what its menu looks like, what each
of its five paths asks you, and where they hand off. It is not a substitute for the role guides,
each pipeline still has its own guide for the domain knowledge (field codes, schema authoring, ODK
app-user management) and real captured output that a short tour like this doesn't cover.

> **A note on the transcripts below.** `eri_do()` is interactive-only, it refuses to run from a
> script, so none of it can execute inside this document the way an ordinary function call could.
> Every menu prompt and message shown here is verified directly against `eri_do()`'s own source
> (`R/wizard.R`), not run live, the same honest convention the [CMR guide](da-cmr-guide.html) and
> [ODK guide](da-odk-guide.html) already use for their non-sandbox-safe steps.

## Why this exists

Earlier, bringing in a monthly report meant memorizing an ordered sequence of function calls and
hand-constructing an Azure path, more to learn than the actual job, which analysts already did by
hand for years. `eri_do()` is the fix: **one guided front door**, decisions instead of syntax. It
never reimplements pipeline logic, every step is one already-tested function
([`eri_upload()`](../reference/eri_upload.html), [`eri_ingest()`](../reference/eri_ingest.html),
[`eri_odk_sync()`](../reference/eri_odk_sync.html), [`eri_onboard_country()`](../reference/eri_onboard_country.html),
and friends), so it is never the *only* way in. Every one of those functions stays fully usable
directly, in a script, in CI, or from the console yourself.

```{=html}
<div class="mermaid">
flowchart TD
    A["eri_do()"] --> B{"What are you trying to do?"}
    B --> C["Bring in this month's CMR"]
    B --> D["Bring in a surveillance dataset"]
    B --> E["Pull in ODK submissions"]
    B --> F["Onboard a new space"]
    B --> G["Review something already staged"]
    C --> H["upload -> stage -> split -> DQ review -> approve"]
    D --> I["archive + DQ + stage (one call) -> approve"]
    E --> J["connect -> register -> sync -> research/raw/"]
    F --> K["dry-run preview -> confirm -> write schema + folders"]
    G --> L["hands off into eri_dq_review()"]
</div>
```

## Before you start {#before-you-start}

```{r library}
library(erifunctions)
```

Just run it from a live R console or RStudio, `eri_do()` refuses to run from `Rscript` or CI (see
[Troubleshooting](troubleshooting.html) if you hit that error unexpectedly):

```{r launch}
eri_do()
```

Always do the same task? Skip the top menu with `flow`:

```{r launch-deep-link}
eri_do("cmr")      # -> straight to the CMR flow
eri_do("ingest")   # -> straight to surveillance ingest
eri_do("odk")      # -> straight to ODK sync
eri_do("onboard")  # -> straight to onboarding
eri_do("review")   # -> straight to the DQ-review shortcut
```

Runs that one flow once and returns, it doesn't loop back into the menu afterward. An unrecognized
name errors immediately, rather than silently falling through to the menu.

## The top-level menu {#top-menu}

Most prompts in the wizard are numbered menus, type the number and press Enter, `0` goes back or
cancels; a few (a country name, a period) are typed instead, blank cancels there. Either way,
nothing is written until you reach a final "Go ahead?" confirmation, see [Cancelling out](#cancelling)
for the full convention. The top menu is five real tasks plus a shortcut into DQ review:

```
── What are you trying to do?

1: Bring this month's country report (CMR) into the system
2: Bring in a surveillance dataset (a CSV/Excel line-list)
3: Pull in ODK survey submissions
4: Onboard a new country, disease, or data type
5: Review & approve something already staged (DQ review)
6: Exit
```

Pick a number, or `6` (or `0`) to leave the wizard entirely. The rest of this guide walks each
path.

## 1. Bring this month's country report (CMR) {#cmr}

```
── Bring this month's country report into the system ────────────────────────────
```

Picks up right where a country's monthly Excel report needs to go: which country (a pick-list of
every registered RB-expansion country, never typed), which file (your OS's native file picker),
then it reads the filename for the reporting month and asks you to confirm rather than typing it:

```
── Which country filed the report?

1: ETH
2: NGA
3: SDN
4: SSD
5: UGA
6: MAD
7: TCD
```

```
ℹ Where is the filled Excel on your computer?
```

An OS file dialog opens next (RStudio's own picker, or the plain OS one). No dialog available
(e.g. a headless session)? It falls back to a typed path instead.

```
── I read the period as 202406 from the file. Is that the reporting month?

1: Yes
2: No
```

If a period wasn't detected, or you say no, it asks directly (`Reporting period, as YYYYMM (e.g.
202406; blank to cancel)`). Then a summary and one confirmation before anything is written:

```
── Ready to bring this in

• Country: UGA
• Month: 202406
• File: uga_cmr_2024_06.xlsx
ℹ I'll upload it, stage it, and split it into per-disease measures. Then we'll review data quality.

── Go ahead?

1: Yes
2: No
```

Say yes, and it uploads, stages, and splits the workbook per disease/measure, all in one run,
auto-detecting whether this country's data also needs mirroring to the legacy pipeline (never
asked, see [`eri_cutover_status()`](../reference/eri_cutover_status.html)). Then it hands off
directly into the same DQ review screen [§5's shortcut](#dq-shortcut) uses, work through any flags,
and it approves for you once everything is clean. If UGA/202406 was already split in an earlier
session, it offers to skip straight to reviewing it instead of starting over.

**The [CMR guide](da-cmr-guide.html)** has the domain knowledge this tour skips: what the field-code
row looks like, why `rblf`/`cmr` is a transitional staging code (not the canonical per-disease
home), and the force-approve escape hatch for a flag that will never resolve cleanly.

## 2. Bring in a surveillance dataset {#ingest}

```
── Bring in a surveillance dataset ───────────────────────────────────────────────
```

Unlike CMR, ingest isn't locked to a registered country list, so country is a typed, validated
code (`sdn`, or the `atlantis` sandbox) rather than a pick-list. Disease, channel (`data_source`),
and measure (`data_type`) are each a pick-list with a typed "Other" escape hatch, since new
combinations are a normal, expected gap, not an error:

```
Which country is this data for? (e.g. sdn; blank to cancel): atlantis
```

```
── Which disease?

1: malaria
2: oncho
3: lf
4: sch
5: sth
6: Other (type it)
```

```
── How did this data arrive?

1: surveillance
2: programmatic
3: research
4: Other (type it)
```

```
── What does each row count?

1: case
2: aggregate
3: treatment
4: mmdp
5: training
6: survey
7: tas
8: prevalence
9: entomology
10: Other (type it)
```

```
ℹ Where is the file on your computer?
```

```
── Ready to bring this in

• Country: ATLANTIS
• Disease: malaria
• Channel: surveillance / Measure: case
• File: atlantis_malaria_2024-01.csv
ℹ I'll archive it, check data quality, and stage it for approval.

── Go ahead?

1: Yes
2: No
```

One call, [`eri_ingest()`](../reference/eri_ingest.html), does the archive + DQ-check + stage in a
single shot, unlike CMR's multi-step upload/stage/split. If it finds open log entries afterward, it
tells you and asks whether to approve anyway, rather than triaging them for you (surveillance's DQ
flags land in the general [`eri_logs()`](../reference/eri_logs.html) backlog, not CMR's per-sheet
shape):

```
! 1 log entry need review before this can be approved.
ℹ Run `eri_logs()` to see them, `eri_dq_flag_resolve()` to triage each flag, then `eri_logs_resolve()` to close the entry out.

── Approve anyway? (only if you've already reviewed this)

1: Yes
2: No
```

Say no (the safer default unless you've genuinely already reviewed it), and it leaves the data
staged rather than approving:

```
ℹ Left staged -- run `eri_do()` again once you've triaged the flags.
```

Otherwise, it asks for the period (free text matched against the staged filename, no fixed
convention the way CMR has one) and approves.

**The [ingest guide](da-ingest-guide.html)** walks the same pipeline by hand, including a full
worked example of a schema catching real problems (an impossible age, an unrecognized district) and
the corrections/flags distinction that drives it.

## 3. Pull in ODK survey submissions {#odk}

```
── Pull in ODK survey submissions ────────────────────────────────────────────────
```

Connects, then turns "which project" and "which form" into pick-lists built from what's actually
visible on your account, never a numeric project id or exact form id typed from memory:

```
── Which ODK project?

1: Uganda
2: testing
```

```
── Which form?

1: ERI Test, River Prospection
```

If this form isn't registered yet, it asks for country (a pick-list of known ERI countries) and
disease, the server URL comes from the connection you already made, never asked again, confirms,
then registers:

```
ℹ This form isn't registered yet -- I need a country and disease to file it under.
```

```
── Which country?

1: DR
2: HT
3: ETH
4: NGA
5: SDN
6: SSD
7: UGA
8: MAD
9: TCD
```

```
── Register this form as uga/oncho?

1: Yes
2: No
```

```
── Sync submissions now?

1: Yes
2: No
```

```
✔ Done. Submissions are synced to research/raw/.
ℹ Next: quality-check and stage this the same way as a surveillance extract -- see the surveillance ingest guide -- then eri_approve() it in.
ℹ Run `eri_do()` again any time to sync new submissions.
```

This is where the wizard **honestly stops**: there is no single stage-then-approve function for
ODK data, the real guide's own next step is a manual [`eri_write()`](../reference/eri_write.html)
after ad hoc quality-checking, so rather than fabricate an approve step the tooling doesn't
support, it hands off here.

**The [ODK guide](da-odk-guide.html)** covers everything the wizard doesn't: standing up a form,
monitoring a survey, and managing who can collect on it.

## 4. Onboard a new country, disease, or data type {#onboard}

```
── Onboard a new country, disease, or data type ──────────────────────────────────
```

Three genuinely different shapes behind one menu, a surveillance space, a CMR space, or an NTD
disease's schemas:

```
── What are you setting up?

1: Surveillance country/disease (case or aggregate reporting)
2: CMR country (monthly Case Management Report)
3: NTD disease (MDA + prevalence schemas, no country folders yet)
```

Picking **surveillance** or **CMR** asks for a country code (typed, since onboarding's whole point
is a country nothing else has registered yet), the full country name, disease (surveillance only),
and a schema-comment language:

```
Full country name (e.g. Uganda; blank to cancel): Atlantis
```

```
── Schema comments in which language?

1: English
2: French
```

Every path previews with a dry run before writing anything, then confirms (surveillance and CMR
each have their own confirm wording; CMR's is shown):

```
── Write this CMR schema template and create the Azure folders?

1: Yes
2: No
```

Already onboarded this country and started filling in the TODOs? The wizard checks for the local
schema file before writing and warns rather than silently overwriting your edits:

```
! Already exists here: 'atlantis_cmr_schema.yaml'

── Overwrite with a fresh template? Any edits already made will be lost.

1: Yes
2: No
```

Picking **NTD disease** skips the Azure folders entirely (local-only, by
[`eri_onboard_disease()`](../reference/eri_onboard_disease.html)'s own design) and asks which
schema(s) you need:

```
── Which schema(s) do you need?

1: Both (MDA + prevalence)
2: MDA only
3: Prevalence only
```

Say yes, and the wizard stops right where it should: the schema template is written (and, for
surveillance/CMR, the Azure folders exist), but filling in the disease-specific columns, validating
the schema, and submitting it via pull request are real domain judgment calls, no wizard should
fabricate those for you.

**The [onboarding guide](da-onboard-guide.html)** picks up from exactly there: what the template
looks like, how to fill in the `TODO` sections, and how to validate and submit it.

## 5. Review & approve something already staged {#dq-shortcut}

The fifth menu item is a shortcut, not its own flow: pick a country and a period, and it hands you
straight into [`eri_dq_review()`](../reference/eri_dq_review.html), the same guided menu the CMR
flow above ends in.

```
── Which country?

1: ETH
2: NGA
3: SDN
4: SSD
5: UGA
6: MAD
7: TCD
```

```
Reporting period, as YYYYMM (e.g. 202406; blank to cancel):
```

**The [DQ review guide](da-dq-review-guide.html)** is the reference for this whole screen, what
each option (fix in source, adjust the schema, mark not important, mark noted, force-approve) does
and why.

## Cancelling out {#cancelling}

Every prompt in the wizard shares one convention: **`0`, or blank on a typed prompt, backs out**.
Nothing is written until you reach a flow's own "Go ahead?"/"Write this...?" confirmation, so you
can explore any menu, back out at any point, and re-run `eri_do()` later with nothing left half
done. A flow that stops partway (declined a confirmation, or the DA cancelled) leaves whatever was
already written in place and tells you so, run `eri_do()` again any time to pick up where you left
off.

## What's next

- **Pick the pipeline for your actual task**: [CMR](da-cmr-guide.html) ·
  [surveillance ingest](da-ingest-guide.html) · [ODK](da-odk-guide.html) ·
  [onboarding](da-onboard-guide.html) · [DQ review](da-dq-review-guide.html).
- **New here?** Start with [Getting started](getting-started.html) and the
  [connections guide](connections-guide.html) before running `eri_do()` for real.
- **Scripting or CI?** `eri_do()` is interactive-only by design, use the scriptable core it calls
  directly: [`eri_upload()`](../reference/eri_upload.html), [`eri_ingest()`](../reference/eri_ingest.html),
  [`eri_odk_sync()`](../reference/eri_odk_sync.html), [`eri_onboard_country()`](../reference/eri_onboard_country.html),
  and the rest of the [pipeline reference](../reference/index.html).

See the [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md) for
the full set of guides.
