---
title: "Triaging and handing back DQ flags, interactively (for data analysts)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Triaging and handing back DQ flags, interactively (for data analysts)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

The [CMR guide](da-cmr-guide.html) and the [QC/feedback guide](da-qc-feedback-guide.html) both walk
the DQ pipeline as a **sequence of functions you call yourself**: check, resolve each flag, close out
the entry, re-run, approve. That's the scriptable core, and it has to exist for automation. But you
shouldn't have to remember that sequence, or which function does which part of it, just to triage a
CMR workbook. **`eri_dq_review()`** is one function, one guided menu, that *is* the whole loop:

```{=html}
<div class="mermaid">
flowchart TD
    A["eri_dq_review(country, period)"] --> B{"Any open flags?"}
    B -- "yes" --> C["Work through flags one at a time<br/>fix in source / adjust schema / mark not important / mark noted"]
    C --> D["Re-run the DQ check"]
    D --> B
    B -- "no" --> E["Nothing outstanding"]
    E --> F["Approve"]
    E --> G["Print report -> eri_dq_export()"]
    G --> E
</div>
```

Everything under the hood is still `eri_cmr_dq_report()`, `eri_dq_flag_resolve()`,
`eri_logs_resolve()`, `eri_split_cmr()`, `eri_approve_cmr()`, and `eri_dq_export()` -- but as a DA
running this interactively, you never call any of them directly. You just answer the menu.

> **The wrapper holds no state of its own.** Every open re-derives current flags fresh from a real,
> logged DQ check against whatever's actually staged -- it never trusts a possibly-stale in-memory
> or cached view. Every triage decision is written through immediately too, not batched. So closing
> your laptop mid-review and running `eri_dq_review()` again later is always safe: it re-checks from
> the real staged data and picks up your prior decisions from the logs. The one thing it doesn't
> remember between separate runs is *which local file you were fixing* -- point it back at your
> working copy (or the original; see [below](#fix-in-source)) if you come back to re-run a check.

## Before you start {#before-you-start}

- `remotes::install_github("thecartercenter/erifunctions")`.
- **Azure access** to the `data` blob (zero-config browser sign-in, see the
  [connections guide](connections-guide.html)).
- `openxlsx2` (already a package dependency) to build the practice workbook below.
- A CMR workbook already staged and split -- see the [CMR guide](da-cmr-guide.html) sections 1-4 for
  a real country. Below, we practice on the built-in **`atlantis`** training sandbox instead, so
  nothing here touches a real program's data.

```{r library}
library(erifunctions)
```

> **What's `atlantis`?** A fictional country the package ships a CMR routing schema and a DQ schema
> for (`load_cmr_schema("atlantis")`, `load_dq_schema("atlantis", "oncho", "programmatic",
> "treatment")`), purely so this exact loop -- split, flag, triage, approve -- can be practiced
> against the real `data` Azure blob without ever touching a real country's namespace.

## 1. Build a practice submission with two real problems {#build}

The bundled `cmr-example.xlsx` (used throughout the [CMR guide](da-cmr-guide.html)) is clean data, so
there's nothing for a DQ check to catch. To have something to actually review, we deliberately
introduce two realistic problems into a copy of it -- the same kind of thing a real submission might
contain: an impossible value, and a locality the schema doesn't recognise yet.

We also drop the `SCH Treatment` sheet: `atlantis` only ships a bundled DQ schema for
`oncho/treatment`, and `eri_approve_cmr()` blocks on **any** routed measure that's never been
DQ-checked (a schema-less measure never gets one) -- the same rule applies to a real, multi-sheet
CMR, where every routed measure needs a schema before the whole workbook can be approved.

```{r build-workbook}
library(openxlsx2)

wb <- wb_load(system.file("extdata", "cmr-example.xlsx", package = "erifunctions"))
wb$remove_worksheet("SCH Treatment")
wb <- wb_add_data(wb, sheet = "RB Treatment", x = -50,
                  start_col = 6, start_row = 6, col_names = FALSE)             # Kampala's treated count
wb <- wb_add_data(wb, sheet = "RB Treatment", x = "Atlantis City",
                  start_col = 4, start_row = 8, col_names = FALSE)             # Soroti's district cell
path <- file.path(tempdir(), "202608_atlantis_demo.xlsx")
wb_save(wb, path, overwrite = TRUE)
```

## 2. Get a workbook with something to review {#split}

`eri_split_cmr()` routes the workbook to `atlantis`'s `oncho/programmatic/treatment` measure;
`eri_cmr_dq_report()` checks it against that measure's schema and combines every flag into one
tibble -- both covered in depth in the [CMR guide](da-cmr-guide.html#dq-check).

```{r split-and-check}
plan  <- eri_split_cmr(path, "atlantis", period = "202608")
flags <- eri_cmr_dq_report("atlantis", "202608", plan = plan)
#> ✔ DQ checks complete: 0 corrections, 2 flags for review.
#> ✔ Logged 2 DQ flags (needs_review).
flags[, c("excel_row", "column", "value", "issue", "status")]
#> # A tibble: 2 × 5
#>   excel_row column   value         issue                                  status
#>       <int> <chr>    <chr>         <chr>                                  <chr>
#> 1         6 treated  -50           Value outside expected range [0, 1000… open
#> 2         8 district Atlantis City Value not in allowed_values list       open
```

Two flags, one measure. Now forget that tibble exists -- from here on, one function.

## 3. Run the review {#review}

```{r review, eval=FALSE}
eri_dq_review("atlantis", "202608")
```

`eri_dq_review()` is **interactive only** -- it refuses to run in a script or from `Rscript`, with a
pointer back to the scriptable core (`eri_cmr_dq_report()`, `eri_dq_flag_resolve()`,
`eri_logs_resolve()`, `eri_approve_cmr()`) for CI/automation. Run it from a live R console or RStudio
session. Below is a real session against the workbook built above, one menu choice at a time -- the
menu prompts themselves come straight from the function; the informational and success messages
between them are real, captured output.

> **Your console will be busier than the transcripts below.** Each DQ check also prints its own
> "renamed column" and "corrections/flags" lines, and you'll likely see
> `! Could not load schema from Azure (Not Found...). Falling back to local.` -- expected here,
> since `atlantis`'s DQ schema only ships inside the package and is never uploaded to Azure. Nothing
> below is broken; the transcripts are trimmed to what changes each step, not a full console dump.

### It opens on the flags {#opens}

```
── DQ review: atlantis / 202608 ────────────────────────────────────────────────
✖ 1 of 1 sheet have open flags (2 flags total)

── RB Treatment (oncho/treatment) -- 2 open ──

 excel_row   column         value                                      issue
         6  treated           -50 Value outside expected range [0, 10000000]
         8 district Atlantis City           Value not in allowed_values list

── 2 open flag(s) across the workbook. What do you want to do?

1: Work through the open flags one at a time
2: Re-run the DQ check
3: Force-approve anyway
4: Print report
5: Exit

```

Two open flags need two different responses, not the same one -- that's the whole reason to work
through them one at a time rather than bulk-resolving.

### Flag 1: a real error -> fix in source {#fix-in-source}

`treated = -50` is impossible -- a genuine data-entry mistake, not a judgement call. Choosing
**"Work through the open flags"**, then **"Fix in source"** for this flag:

```
── Flag 1/2: RB Treatment row 6
treated: "-50" -- Value outside expected range [0, 10000000]
── What do you want to do with this flag?

1: Fix in source (open/copy the workbook)
2: Adjust the schema (alias, allowed value, range...)
3: Mark not important
4: Mark noted
5: Skip to the next flag

Path to the local source workbook for this period (or a '_fixed' copy you've already started, if you have one): <you type the path here>
✔ Made a working copy: '202608_atlantis_demo_fixed.xlsx' (original preserved)
ℹ Fix treated on Excel row 6 in the "RB Treatment" sheet.
ℹ Issue: Value outside expected range [0, 10000000]
ℹ Open this file to review/edit: '.../202608_atlantis_demo_fixed.xlsx'
ℹ When you're done with this and any other fixes, choose "Re-run the DQ check" from the main menu -- it re-splits '202608_atlantis_demo_fixed.xlsx' and re-checks.
```

The first time you fix anything in a session, `eri_dq_review()` forks a `_fixed` working copy before
touching it -- the file as submitted is never edited in place. Open that copy, correct the value,
save it. (Picking this back up in a fresh R session? Point it at that same `_fixed` file, or the
original again -- it detects the `_fixed` sibling and reuses it rather than forking a second one.)

> In RStudio, the working copy opens directly in the editor instead of printing an "open this file"
> message -- `eri_dq_review()` uses `rstudioapi` when it's available.

### Flag 2: not wrong, just not catalogued yet -> defer it {#defer}

`"Atlantis City"` isn't a typo -- it's a real locality the schema's `allowed_values` list hasn't
caught up to yet. There's nothing to *fix*. Choose **"Skip to the next flag"** for now (we'll come
back to it once the actual error above is out of the way):

```
── Flag 2/2: RB Treatment row 8
district: "Atlantis City" -- Value not in allowed_values list
```

### Re-run: only what changed gets re-checked {#rerun}

Back at the main menu (still 2 open -- fixing the workbook doesn't retroactively change the logged
flags), choose **"Re-run the DQ check"**:

```
── Re-split '202608_atlantis_demo_fixed.xlsx' and re-check just what it routes to?

1: Yes
2: No -- cancel

```

```
ℹ Superseded a prior staged file for this period: '202608_atlantis_demo_rb_treatment.parquet'
✔ "RB Treatment" -> 'atlantis/oncho/programmatic/treatment/staged/202608_atlantis_demo_fixed_rb_treatment.parquet'
✔ DQ checks complete: 0 corrections, 1 flag for review.

── DQ review: atlantis / 202608 ────────────────────────────────────────────────
✖ 1 of 1 sheet have open flags (1 flag total)

── RB Treatment (oncho/treatment) -- 1 open ──

 excel_row   column         value                            issue
         8 district Atlantis City Value not in allowed_values list
```

The `treated` flag is gone -- the underlying value is fixed, so it never comes back. The
`"Atlantis City"` flag reappears exactly as it was, because it's genuinely still there. If your
workbook routes several measures, only the ones you actually re-split get re-checked; every other
measure's in-session decisions are left alone (see the [CMR guide](da-cmr-guide.html#dq-check) for a
worked multi-measure example).

### Flag 2, revisited: mark it noted {#mark-noted}

One flag left. **"Work through the open flags"** again, then **"Mark noted"** this time, with an
explanation:

```
── Flag 1/1: RB Treatment row 8
district: "Atlantis City" -- Value not in allowed_values list
Note (optional): <you type your explanation here>
✔ Flag 1 in '...dq_flags_202608.yaml' marked "noted".
```

We typed: *"Confirmed with the country lead: Atlantis City is a real, if not-yet-catalogued,
locality -- flagging for the schema maintainer to add it to the allowed list."* That note travels
with the flag from here on -- into the log, and into the handback file below.

### Nothing outstanding -> print the report {#print-report}

```
── DQ review: atlantis / 202608 ────────────────────────────────────────────────
✔ 1 sheet checked -- all clean.
── Nothing outstanding. What next?

1: Approve
2: Print report
3: Exit

```

"All clean" means every flag has been triaged, not that the data is now flawless -- one flag is just
`noted` rather than `open`. Choose **"Print report"**:

```
── RB Treatment
 excel_row   column         value                            issue status
         8 district Atlantis City Value not in allowed_values list  noted
                                                                                                                                                           note
 Confirmed with the country lead: Atlantis City is a real, if not-yet-catalogued, locality -- flagging for the schema maintainer to add it to the allowed list.

✔ DQ report (1 flag · 0 open) written to '.../dq-report-atlantis-202608-2026-07-12.html'.
```

A quick console eyeball first -- a long `note` doesn't truncate, it just pushes the table wide, as it
does here -- then the handback file. `eri_dq_review()` calls [`eri_dq_export()`](../reference/eri_dq_export.html) for you here -- see the
[QC/feedback guide](da-qc-feedback-guide.html#export) for what that function does on its own. Open
the file: a self-contained page, organised by sheet, with the `noted` status shown as a chip and the
note text in its own column -- exactly what you'd forward to the country or paste into Teams, with
zero reformatting.

### Approve {#approve}

Back at the same "Nothing outstanding" menu (print report loops back to it, it doesn't exit), choose
**"Approve"**:

```
✔ Catalog: registered '202608_atlantis_demo_fixed_rb_treatment.parquet'.
✔ Approved: '202608_atlantis_demo_fixed_rb_treatment.parquet'
✔ Approval log: 'atlantis/oncho/programmatic/treatment/processed/202608_approval_log.yaml'
── ✔ Approved "202608" ─────────────────────────────────────────────────────────
Dataset: atlantis / oncho / programmatic / treatment
Files: 1 moved to processed
Approver: NishantKishore (unverified)
Location: atlantis/oncho/programmatic/treatment/processed
ℹ Operation log: 'atlantis/oncho/programmatic/treatment/logs/20260712_014108_eri_approve_202608.yaml'
ℹ Operation log: 'atlantis/rblf/cmr/logs/20260712_014108_eri_approve_cmr_202608.yaml'
✔ Approved 1 measure for "atlantis" / "202608".
```

> `Approver` shows as "(unverified)" here because this sandbox session has no `ERI_ANALYST_ID` set --
> on a normal analyst machine it carries your verified identity instead (see the
> [onboarding guide](onboarding.html)).

That's the whole session: two flags, two different resolutions, a handback file, one approval --
without calling `eri_dq_flag_resolve()`, `eri_logs_resolve()`, or `eri_approve_cmr()` yourself even
once.

## Cleaning up {#cleanup}

Everything above lives under `atlantis`'s own namespace, so it's safe to leave -- but if you were
following along and want to tidy up the sandbox behind you, three things were created: the Azure-side
staged/log/processed files, the local practice workbook (`path`) and its `_fixed` working copy
(forked in [Flag 1](#fix-in-source)), and the exported HTML report (`eri_dq_export()` writes to your
working directory by default, not a temp folder -- see the
[QC/feedback guide](da-qc-feedback-guide.html#export) -- so it won't clean itself up):

```{r cleanup}
con <- get_azure_storage_connection(storage_name = Sys.getenv("ERIFUNCTIONS_DATA_STORAGE_NAME", "data"))
eri_catalog_remove(
  "atlantis/oncho/programmatic/treatment/processed/202608_atlantis_demo_fixed_rb_treatment.parquet",
  data_con = con
)
all_files <- AzureStor::list_storage_files(con, "atlantis", recursive = TRUE)
mine      <- all_files$name[!all_files$isdir & grepl("202608", all_files$name)]
for (f in mine) AzureStor::delete_storage_file(con, f, confirm = FALSE)

unlink(path)                                                    # the original practice workbook
unlink(sub("\\.xlsx$", "_fixed.xlsx", path))                     # its "fix in source" working copy
unlink(list.files(getwd(), pattern = "^dq-report-atlantis-202608-.*\\.html$", full.names = TRUE))
```

Real submissions are not disposable like this practice run -- never delete a real country's staged,
processed, or log files this way.

## What's next

- [`eri_dq_export()`](../reference/eri_dq_export.html) on its own, and its markdown output, in the
  [QC/feedback guide](da-qc-feedback-guide.html#export).
- The scriptable core this orchestrates, and a worked multi-measure example, in the
  [CMR guide](da-cmr-guide.html#dq-check).
- Scanning the *logged* state of the whole team's backlog (not just one session's tibble) with
  `eri_logs()`, the [log-triage guide](da-logs-guide.html).
- Submitting a schema fix for a maintainer to fold in permanently (rather than a one-off note) via
  `eri_dq_schema_submit()` -- offered automatically at the end of a session if you used "Adjust the
  schema" on any flag.
