---
title: "Ingesting a surveillance dataset: raw to approved (for data analysts)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ingesting a surveillance dataset: raw to approved (for data analysts)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This is a hands-on, **start-to-finish walkthrough** of getting a surveillance dataset into the
Carter Center data system, the daily work of a **Data Analyst (DA)**. It is written for domain
experts who are *not* software developers. Work through it once, in order, and you will have done
every step a real dataset needs: store the raw file, quality-check it, stage it, and **approve** it
so the rest of the team can trust it.

So you can practise the whole thing safely, we will invent a **make-believe country**, *Atlantis*,
and make up some malaria surveillance data for it. Nothing here touches real country data, and the
very last section, [**Clean up**](#clean-up), deletes everything you created, so you can run the
full loop end-to-end and leave no trace.

> **You can run this for real.** Every command below works against the live Azure system. Because
> *Atlantis* is a sandbox of your own, you can ingest, break, fix, and approve data without
> affecting anyone, then wipe it.

## The golden rule

> **Data moves through three layers, and nothing becomes official without your sign-off.** Every
> dataset travels `raw/` &rarr; `staged/` &rarr; `processed/`. `raw/` is the file exactly as it
> arrived. `staged/` is the cleaned, quality-checked version awaiting review. `processed/` is the
> **canonical data the whole team builds on**, and a file only gets there when *you* call
> [`eri_approve()`](../reference/eri_approve.html). That approval is the human gate.

```{=html}
<div class="mermaid">
flowchart TD
    A["Surveillance file arrives"] --> B["Store as-received in raw/"]
    B --> C["Run DQ checks against a schema"]
    C --> D["Fix flags, write cleaned data to staged/"]
    D --> E["eri_approve() = your sign-off"]
    E --> F["processed/ + catalog entry"]
    F -.->|next reporting period| C
    F --> Z["Clean up the sandbox"]
</div>
```

## Before you start {#before-you-start}

You need two things, once:

1. **R and RStudio** installed on your laptop.
2. **The package**, installed from the Carter Center repository:
   ```{r install}
   install.packages("remotes")
   remotes::install_github("thecartercenter/erifunctions")
   ```

**Azure access** needs no configuration. The first time a command needs Azure, your browser opens
and you sign in with your Carter Center account; the sign-in is remembered for the rest of the
session. (No keys, no environment variables, see `?get_azure_storage_connection`.)

`erifunctions` narrates what it is doing. Load it and leave the verbosity at its default `"full"`
for this walkthrough, that is the step-by-step output shown in the blocks below:

```{r verbosity}
library(erifunctions)

# The outputs shown in this guide are at the default "full" verbosity.
# Once you know the steps, eri_verbosity("quiet") trims them to headline
# results, warnings, and errors only.
```

## 1. Connect to the data blob and name your sandbox {#sandbox}

The surveillance pipeline lives in the **`data` blob**. The reading and writing functions
(`eri_read()`, `eri_write()`, `eri_upload()`) talk to your *default* connection, so we make one
pointed at `data` and pass it explicitly everywhere below, that way every step targets the right
place:

```{r connect}
data_con <- get_azure_storage_connection(storage_name = "data")
```

Every dataset is identified by five coordinates ([ADR-0012](https://github.com/thecartercenter/erifunctions/blob/main/docs/adr/0012-source-measure-data-model.md)):
**country**, **disease**, **data_source** (the *channel*, how the data arrives), **data_type** (the
*measure*, what it captures), and **layer**. The key idea is that the channel is **separate** from the
measure: `surveillance` data can be a `case` line-list in one country and an `aggregate` weekly count in
another. Our extract is a line-list, one row per case, so its measure is `case`. Never hand-type the
folder paths, let [`eri_data_path()`](../reference/eri_data_path.html) build the canonical path for
you, so every step agrees on where things go:

```{r data-path}
# Our make-believe country and its malaria surveillance data.
country     <- "atlantis"
disease     <- "malaria"
data_source <- "surveillance"   # the channel
data_type   <- "case"           # the measure (one row per case)

eri_data_path(country, disease, data_source, data_type, "raw")
#> [1] "atlantis/malaria/surveillance/case/raw"

eri_data_path(country, disease, data_source, data_type, "staged")
#> [1] "atlantis/malaria/surveillance/case/staged"
```

`data_source` and `data_type` are **extensible**, `eri_data_model()` lists the known values, and an
unregistered one *warns* rather than blocks, while `layer` is one of `"raw"`, `"staged"`,
`"processed"`. **Country and disease are free text**, which is exactly why we can stand up an *atlantis*
sandbox next to the real countries without disturbing them.

Create the two folders we will write into (the third, `processed/`, is created for us at approval):

```{r make-dirs}
eri_dir_create(eri_data_path(country, disease, data_source, data_type, "raw"),    azcontainer = data_con)
eri_dir_create(eri_data_path(country, disease, data_source, data_type, "staged"), azcontainer = data_con)
#> ✔ Directory created!
```

## 2. Simulate an as-received extract and store it in `raw/` {#raw}

In real life a surveillance officer emails you a spreadsheet. Here we *make one up*, a tiny
line-list of malaria cases, one row per case. Notice it is deliberately a little messy, the way real
data always is: some sex values are in Spanish (`Masculino`/`Femenino`/`M`), and a couple of
district names are mis-cased or use an old spelling (`coralia`, `Nerei City`):

```{r make-extract1, eval=TRUE}
extract1 <- data.frame(
  Year     = rep(2024, 10),
  EpiWeek  = 1:10,
  District = c("Coralia", "Nerei", "coralia", "Tritonis", "Marisa",
               "Nerei City", "Coralia", "Nerei", "Tritonis", "Marisa"),
  Sex      = c("Male", "Female", "Masculino", "Femenino", "M",
               "Female", "Male", "Female", "Masculino", "Male"),
  Age      = c(34, 27, 5, 60, 41, 19, 73, 8, 52, 30),
  stringsAsFactors = FALSE
)
head(extract1)
```

The **first thing** you do with any file is preserve it untouched, so there is always a record of
what actually arrived. Write it to `raw/`:

```{r write-raw1}
raw_path1 <- eri_data_path(country, disease, data_source, data_type, "raw",
                           "atlantis_malaria_2024-01.csv")
eri_write(extract1, raw_path1, azcontainer = data_con)
```

`raw/` is your archive of the source. From here on we work toward `staged/`, but the original is
safe if you ever need to start over.

## 3. Author a data-quality schema {#schema}

How does `erifunctions` know that `Masculino` should become `Male`, or that `Age` cannot be 250? You
tell it, once, in a **DQ schema**, a small YAML file describing each column: its type, its allowed
range or allowed values, known fixes, and translations. The schema is the contract every future
extract for this dataset is checked against.

Open a text editor and save the following as **`atlantis_malaria_surveillance_case.yaml`** in your
working directory, the filename is the four-part identity `{country}_{disease}_{data_source}_{data_type}`,
which is how `load_dq_schema()` finds it. (It is intentionally short, a real schema like
`dr_malaria_surveillance_case.yaml` has more columns, but the *shape* is exactly this.)

> **One source, many measures.** If you stood up this space with the
> [onboarding guide](da-onboard-guide.html), its scaffold defaults to the **`aggregate`** measure
> (weekly counts), `atlantis_malaria_surveillance_aggregate.yaml`. Our extract here is a line-list, so
> we author the **`case`** measure alongside it. That is the whole point of separating the channel from
> the measure: the same `surveillance` source can hold both `case` and `aggregate` data. Run
> `eri_data_model()` any time to see the known measures.

```yaml
country: atlantis
disease: malaria
data_source: surveillance
data_type: case
language: en
time_grain: case

temporal:
  year_col: Year
  period_col: EpiWeek

preprocessing:
  - remove_smart_quotes
  - drop_rows_missing_year

columns:
  Year:
    required: true
    type: numeric
    aliases: [year, YEAR, Año]
    range: [2000, 2035]

  EpiWeek:
    required: true
    type: numeric
    aliases: [week, Week, Semana]
    range: [1, 53]

  District:
    required: true
    type: categorical
    aliases: [district, Distrito]
    allowed_values: [Coralia, Nerei, Tritonis, Marisa]
    corrections:
      coralia: Coralia
      CORALIA: Coralia
      "Nerei City": Nerei

  Sex:
    required: true
    type: categorical
    aliases: [sex, SEX, Sexo]
    allowed_values: [Male, Female]
    translations:
      Masculino: Male
      Femenino: Female
      M: Male
      F: Female

  Age:
    required: true
    type: numeric
    aliases: [age, AGE, Edad]
    range: [0, 120]
```

Read it slowly, each block is a rule:

| Block | What it does | DQ result |
|---|---|---|
| `aliases` | accepts other column names (e.g. a Spanish `Edad`) and renames them | quietly renamed |
| `type` + `range` | coerces to the type, then flags values outside `[min, max]` | **flag** for you to review |
| `allowed_values` | flags anything not on the list | **flag** for you to review |
| `corrections` | auto-replaces known bad spellings with the canonical value | **auto-fixed**, logged |
| `translations` | auto-maps synonyms (e.g. `Masculino` &rarr; `Male`) | **auto-fixed**, logged |

The distinction that matters: **corrections and translations are applied for you and recorded;
flags are problems only *you* can resolve.**

Now publish the schema to Azure so [`load_dq_schema()`](../reference/load_dq_schema.html) can find
it. It looks for `schemas/{country}_{disease}_{data_source}_{data_type}.yaml`, so the filename matters:

```{r upload-schema}
eri_upload("atlantis_malaria_surveillance_case.yaml",
           "schemas/atlantis_malaria_surveillance_case.yaml", azcontainer = data_con)
```

## 4. Run the data-quality checks {#dq}

Load your raw file back from `raw/`, load the schema, and run the checks. `run_dq_checks()` applies
every rule in the schema and hands back the cleaned data plus two ledgers, what it *fixed*, and
what it wants *you* to look at:

```{r dq1}
raw1    <- eri_read(raw_path1, azcontainer = data_con)
schema  <- load_dq_schema("atlantis", "malaria", "surveillance", "case", azcontainer = data_con)

result1 <- run_dq_checks(raw1, schema)
#> ✔ DQ checks complete: 6 corrections, 0 flags for review.
```

Print the result for the full report:

```{r dq1-report}
result1
#> ── Data Quality Report ──────────────────────────────────────────────
#> Shape: 10 rows x 5 columns
#>
#> ── Automated Corrections (6 total) ──
#> • "Sex": 4 corrections (translation)
#> • "District": 2 corrections (correction)
#>
#> ── Flags Requiring Review (0 total) ──
#> ✔ No flags.
```

Six things were fixed automatically and **nothing needs your review**, a clean extract. You can
always see exactly *what* was changed in the `$log` ledger:

```{r dq1-log}
result1$log
#> # A tibble: 6 × 6
#>     row column   original_value corrected_value rule        action
#>   <int> <chr>    <chr>          <chr>           <chr>       <chr>
#> 1     3 Sex      Masculino      Male            translation corrected
#> 2     9 Sex      Masculino      Male            translation corrected
#> 3     4 Sex      Femenino       Female          translation corrected
#> 4     5 Sex      M              Male            translation corrected
#> 5     3 District coralia        Coralia         correction  corrected
#> 6     6 District Nerei City     Nerei           correction  corrected
```

The cleaned data lives in `result1$data`, Spanish sexes mapped to `Male`/`Female`, district names
canonicalised. **That** is what we stage.

## 5. Write the cleaned data to `staged/` {#stage}

`staged/` holds DQ-checked data that is awaiting your approval. We store it as **parquet** (compact,
typed, fast). Give the file a name that contains the reporting period, you will use that period to
approve it in the next step:

```{r stage1}
staged_path1 <- eri_data_path(country, disease, data_source, data_type, "staged",
                              "atlantis_malaria_2024-01.parquet")
eri_write(result1$data, staged_path1, azcontainer = data_con)
```

Nothing is official yet. The data is checked and parked, but it will not appear as canonical until
you sign off.

## 6. Approve it into `processed/` {#approve}

This is the gate. [`eri_approve()`](../reference/eri_approve.html) finds every staged file whose
name contains the period you give it, moves it to `processed/`, writes a human-readable **approval
log** (who approved what, and when), and **registers the file in the data catalog** so teammates can
discover it:

```{r approve1}
eri_approve(country, disease, data_source, "2024-01", data_type = data_type, azcontainer = data_con)
#> ✔ Catalog: registered atlantis_malaria_2024-01.parquet.
#> ✔ Approved: atlantis_malaria_2024-01.parquet
#> ✔ Approval log: atlantis/malaria/surveillance/case/processed/2024-01_approval_log.yaml
#> ── ✔ Approved "2024-01" ─────────────────────────────────
#> Dataset: atlantis / malaria / surveillance / case
#> Files: 1 moved to processed
#> Approver: your.name
#> Location: atlantis/malaria/surveillance/case/processed
#> ℹ Operation log: atlantis/malaria/surveillance/case/logs/20240115_142233_eri_approve_2024-01.yaml
```

Passing `data_type` records the **measure** on the catalog entry. (Omit it and `eri_approve()` files the
data at the channel level with `data_type = NA`, and says so once, fine for channel-only data, but here
our line-list is `case`.)

The catalog is the team's index of canonical data. Confirm your dataset is now in it:

```{r catalog1}
eri_catalog_query(country = "atlantis", data_con = data_con)
#> # A tibble: 1 × 13
#>   path                        country disease data_source data_type layer  period  …
#>   <chr>                       <chr>   <chr>   <chr>       <chr>     <chr>  <chr>
#> 1 atlantis/malaria/surveill…  atlantis malaria surveillance case      proce… 2024-01
```

And read the canonical file straight back, exactly as any teammate would:

```{r read-processed}
processed_path1 <- eri_data_path(country, disease, data_source, data_type, "processed",
                                 "atlantis_malaria_2024-01.parquet")
official <- eri_read(processed_path1, azcontainer = data_con)
nrow(official)   # 10 approved cases
```

That is the whole loop: **raw &rarr; DQ &rarr; staged &rarr; approved.**

## 7. A second extract arrives, this time with errors {#second-extract}

The next reporting file lands, and real data is rarely clean. This one has two genuine problems an
analyst must judge: an `Age` of **250** (impossible), and a district **`Atlantica`** that is not one
of our four (a typo, or a new district the schema does not know yet):

```{r make-extract2, eval=TRUE}
extract2 <- data.frame(
  Year     = rep(2024, 10),
  EpiWeek  = 11:20,
  District = c("Coralia", "Atlantica", "Nerei", "Tritonis", "Marisa",
               "coralia", "Nerei", "Coralia", "Tritonis", "Marisa"),
  Sex      = c("Male", "Female", "Masculino", "Female", "M",
               "Female", "Male", "Femenino", "Male", "Female"),
  Age      = c(34, 27, 250, 60, 41, 19, 73, 8, 52, 30),
  stringsAsFactors = FALSE
)
```

Archive it to `raw/`, then run the same checks:

```{r dq2}
raw_path2 <- eri_data_path(country, disease, data_source, data_type, "raw",
                           "atlantis_malaria_2024-02.csv")
eri_write(extract2, raw_path2, azcontainer = data_con)

raw2    <- eri_read(raw_path2, azcontainer = data_con)
result2 <- run_dq_checks(raw2, schema)
#> ✔ DQ checks complete: 4 corrections, 2 flags for review.

result2
#> ── Data Quality Report ──────────────────────────────────────────────
#> Shape: 10 rows x 5 columns
#>
#> ── Automated Corrections (4 total) ──
#> • "Sex": 3 corrections (translation)
#> • "District": 1 correction (correction)
#>
#> ── Flags Requiring Review (2 total) ──
#> ! Value not in allowed_values list: 1 row [District] (e.g. Atlantica (row 2))
#> ! Value outside expected range [0, 120]: 1 row [Age] (e.g. 250 (row 3))
#>
#> ℹ See `result$flags` for the full row-level detail.
```

This time there are **two flags**. The schema fixed what it safely could, but it will not guess about
an age of 250 or an unknown district, those are *your* call. Look at exactly which rows:

```{r dq2-flags}
result2$flags
#> # A tibble: 2 × 4
#>     row column   value     issue
#>   <int> <chr>    <chr>     <chr>
#> 1     3 Age      250       Value outside expected range [0, 120]
#> 2     2 District Atlantica Value not in allowed_values list
```

You investigate (call the surveillance officer, check the source form) and decide: the age was a
data-entry slip for **25**, and `Atlantica` is really **Marisa**. Fix them in the cleaned data, then
re-run the checks to confirm a clean bill of health:

```{r fix2}
fixed <- result2$data
fixed$Age[3]      <- 25
fixed$District[2] <- "Marisa"

result2b <- run_dq_checks(fixed, schema)
#> ✔ DQ checks complete: 0 corrections, 0 flags for review.
```

> **Why not just edit the schema?** If `Atlantica` were a *real* new district, the right fix is to
> add it to `allowed_values` and re-upload the schema, then every future extract accepts it. Use a
> correction/translation for genuine fixes; use a flag-then-edit only for one-off data errors like
> the impossible age.

Now stage and approve the corrected second period, exactly as before:

```{r stage-approve2}
staged_path2 <- eri_data_path(country, disease, data_source, data_type, "staged",
                              "atlantis_malaria_2024-02.parquet")
eri_write(result2b$data, staged_path2, azcontainer = data_con)

eri_approve(country, disease, data_source, "2024-02", data_type = data_type, azcontainer = data_con)
#> ✔ Catalog: registered atlantis_malaria_2024-02.parquet.
#> ✔ Approved: atlantis_malaria_2024-02.parquet
#> ── ✔ Approved "2024-02" ─────────────────────────────────
#> Dataset: atlantis / malaria / surveillance / case
#> Files: 1 moved to processed
#> Approver: your.name
#> Location: atlantis/malaria/surveillance/case/processed
```

You now have **two approved periods** in the catalog for *atlantis*, each with its own approval log
recording that you signed off and when.

> **Going further: cross-period anomalies.** Range and allowed-value checks catch bad *cells*. To
> catch a suspicious *trend*, a district whose case count suddenly triples, see
> [`add_anomaly_pct_change()`](../reference/add_anomaly_pct_change.html), which you can chain onto a
> `run_dq_checks()` result once you have several periods stacked together.

## 8. Clean up {#clean-up}

You ran this for practice, so remove everything it created and leave no trace. This deletes the whole
*atlantis* sandbox (all three layers and the logs), the schema, and the catalog entries:

```{r cleanup}
# Delete the entire sandbox namespace (raw + staged + processed + logs).
# eri_dir_delete() removes a non-empty directory recursively in one call; using
# the erifunctions helpers (not raw AzureStor) keeps every step in the package.
eri_dir_delete("atlantis", azcontainer = data_con)

# Delete the schema you published.
eri_delete("schemas/atlantis_malaria_surveillance_case.yaml", azcontainer = data_con)

# Remove the two catalog entries (the catalog is shared, so tidy up after yourself).
eri_catalog_remove("atlantis/malaria/surveillance/case/processed/atlantis_malaria_2024-01.parquet",
                   data_con = data_con)
eri_catalog_remove("atlantis/malaria/surveillance/case/processed/atlantis_malaria_2024-02.parquet",
                   data_con = data_con)
#> ✔ Catalog: removed atlantis_malaria_2024-01.parquet.
#> ✔ Catalog: removed atlantis_malaria_2024-02.parquet.
```

Confirm nothing of yours remains, then delete the local files:

```{r cleanup-verify}
eri_catalog_query(country = "atlantis", data_con = data_con)
#> No catalog entries match the specified filters.

unlink("atlantis_malaria_surveillance_case.yaml")   # the local schema file
```

> **Why we could delete freely here:** *Atlantis* is invented and its data is fake. **Real Carter
> Center surveillance data is different**, approved country data must never be deleted casually,
> copied off Azure, or committed to git. The discipline this guide builds, raw kept untouched, DQ
> before staging, and a human approval gate before anything becomes canonical, is exactly what keeps
> protected data trustworthy. When in doubt, ask before you delete.

## What's next

You have now done the full DA ingest loop: store raw, quality-check against a schema, fix what only a
human can, stage, and **approve** into the canonical layer, twice, including a messy extract.

One call, [`eri_ingest()`](../reference/eri_ingest.html), does the read-and-DQ-and-stage steps
(§2, §4, §5) in a single shot, on **any** data including the sandbox you just built (it takes the same
`country` / `disease` / `data_source` / `data_type` you have been using). We did each step by hand here
so you could *see* every layer; once you are comfortable, `eri_ingest()` is the fast path, and
`eri_approve()`: your sign-off, is always the same human gate at the end. *(For the legacy `hsp-mal`
cutover, passing `mirror_pipeline = "hsp-mal"` also mirrors the output to the old `projects` blob, off
by default.)*

This is one of a planned set of short, task-oriented guides, **one for each common job** an analyst
or epidemiologist does. See [the guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md)
for what exists today and what is coming, and the [reference](../reference/index.html) for the full
menu of functions grouped by purpose.
