---
title: "Triaging the error & data-quality log backlog (for data analysts)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Triaging the error & data-quality log backlog (for data analysts)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Every operation `erifunctions` runs — an ingest, an approval, an ODK sync — **leaves a log** in
Azure. When something fails, or a data-quality check raises issues, that log is the record. This guide
is for a **Data Analyst** working the **backlog**: finding what failed or needs review with
[`eri_logs()`](../reference/eri_logs.html), and closing items out with
[`eri_logs_resolve()`](../reference/eri_logs_resolve.html).

Because the logs live in Azure, the backlog is **shared** — if a teammate is out, you can run
`eri_logs()`, see exactly what broke, and pick up where they left off. We will practise on the
**`atlantis`** sandbox from the [ingest](da-ingest-guide.html) and [onboarding](da-onboard-guide.html)
guides.

## The golden rule

> **Nothing fails silently, and nothing stays open by accident.** Operations write structured logs as
> they go; data-quality flags are persisted with [`eri_dq_log()`](../reference/eri_dq_log.html) (and
> `eri_ingest()` does it for you). `eri_logs()` turns that pile into a triage list, and
> `eri_logs_resolve()` marks an item handled — with *who* handled it and *when* — so it drops off the
> open backlog without ever erasing the record.

```{=html}
<div class="mermaid">
flowchart TD
    A["An operation runs (ingest / approve / sync)"] --> B["Writes a log to …/logs/"]
    C["run_dq_checks() flags issues"] --> D["eri_dq_log() persists them"]
    B --> E["eri_logs() — the backlog"]
    D --> E
    E --> F["eri_logs_resolve() — mark handled"]
    F --> G["Drops off the open backlog (record kept)"]
</div>
```

## Before you start {#before-you-start}

Install the package and sign in to Azure (zero-config browser; see the
[connections guide](connections-guide.html)).

```{r library}
library(erifunctions)
```

## 1. Where logs come from {#sources}

Two kinds of log end up in `{country}/{disease}/{data_type}/logs/`:

- **Operation logs** — written automatically by `eri_ingest()`, `eri_approve()`, `eri_stage()`,
  `eri_odk_sync()`, and friends. Each records the operation, who ran it, its parameters, and — if it
  failed — the **error**.
- **Data-quality logs** — the `$flags` from [`run_dq_checks()`](../reference/run_dq_checks.html) are
  only in your R session until you persist them. `eri_dq_log()` writes them to the same place (and
  `eri_ingest()` calls it automatically), so a data-quality issue someone hit last month is still
  discoverable today.

Let's make one of each in the sandbox.

### Persist a data-quality check

Picking up from the [ingest guide](da-ingest-guide.html), you have a `run_dq_checks()` result with a
couple of flags. Persist it:

```{r dq-log}
# result <- run_dq_checks(raw, schema)   # from the ingest guide
eri_dq_log(result, "atlantis", "malaria", "surveillance", period = "2024-01")
#> ℹ Operation log: atlantis/malaria/surveillance/logs/20260626_093000_dq_flags_2024-01.yaml
#> ✔ Logged 2 DQ flags (needs_review).
```

### Cause a failure

Now a common slip: you try to approve a period before the file is staged. It errors — **and leaves an
error log** behind:

```{r approve-fail}
eri_approve("atlantis", "malaria", "surveillance", "2024-02")
#> ℹ Operation log: atlantis/malaria/surveillance/logs/20260626_100000_eri_approve_2024-02.yaml
#> Error: No staged files found matching "2024-02" in atlantis/malaria/surveillance/staged.
```

## 2. See the backlog {#backlog}

`eri_logs()` reads those logs and hands you a triage list, newest first. Scope it to a dataset (fast),
or call it bare for a system-wide view:

```{r logs}
eri_logs("atlantis", "malaria", "surveillance")
#> 2 logs (2 needing attention).
#> # A tibble: 2 × 15
#>   timestamp            operation   status       period  summary           analyst …
#>   <chr>                <chr>       <chr>        <chr>   <chr>             <chr>
#> 1 2026-06-26T10:00:00Z eri_approve error        2024-02 No staged files…  da.one
#> 2 2026-06-26T09:30:00Z dq_flags    needs_review 2024-01 2 flags           da.one
```

(The tibble also carries `log_path`, `n_issues`, and the `handled*` columns — shown trimmed here.)
Filter to just the failures:

```{r logs-error}
eri_logs("atlantis", "malaria", "surveillance", status = "error")
#> 1 log (1 needing attention).
#> # A tibble: 1 × 15
#>   operation   status period  summary
#>   <chr>       <chr>  <chr>   <chr>
#> 1 eri_approve error  2024-02 No staged files found matching '2024-02'.
```

`status = "needs_review"` narrows to data-quality items instead; `operation =`, `analyst =`, and
`since =` filter the rest. The tibble also carries `data_source` (the channel) and `data_type` (the
measure, when the data was approved with one). Leaving `country`/`disease`/`data_source`/`data_type`
out scans the whole data blob for a **system-wide backlog** (scoping is faster).

## 3. Resolve an item {#resolve}

You investigate the failure, stage the file, and re-approve it successfully. Now close out the original
error so it leaves the open backlog — with a note for whoever reads it next:

```{r resolve}
backlog <- eri_logs("atlantis", "malaria", "surveillance", status = "error")

eri_logs_resolve(
  backlog$log_path[1],
  note = "Staged the file, then re-approved 2024-02."
)
#> ✔ Marked 20260626_100000_eri_approve_2024-02.yaml handled.
```

Re-list, and the error is gone from the open backlog — only the data-quality item remains:

```{r relist}
eri_logs("atlantis", "malaria", "surveillance")
#> 1 log (1 needing attention).
#> # A tibble: 1 × 15
#>   operation status       period  summary
#>   <chr>     <chr>        <chr>   <chr>
#> 1 dq_flags  needs_review 2024-01 2 flags
```

Resolving never deletes the log — it adds a `triage` record. Pass `include_handled = TRUE` to see the
closed item, with who handled it and your note:

```{r include-handled}
eri_logs("atlantis", "malaria", "surveillance",
         status = "error", include_handled = TRUE)[, c("operation", "handled", "handled_by", "handled_at")]
#> # A tibble: 1 × 4
#>   operation   handled handled_by handled_at
#>   <chr>       <lgl>   <chr>      <chr>
#> 1 eri_approve TRUE    da.one     2026-06-26T10:15:00Z
```

## 4. When a teammate steps in {#teammate}

This is the point of a *shared* backlog. The logs live in Azure, not on anyone's laptop — so when an
analyst is away and something needs attention, a colleague can run one command to see the whole
picture:

```{r teammate}
# Anyone on the team, across every dataset:
eri_logs(status = "error")
```

They see the same failures, the same error messages, and — for anything already handled — *who* dealt
with it and the note they left. No "what was going on with the DR upload last week?" guesswork.

## 5. Clean up {#clean-up}

If you ran this in the sandbox, remove the `atlantis` namespace (this deletes its logs too):

```{r cleanup}
con <- get_azure_storage_connection(storage_name = "data")
# eri_dir_delete() removes the namespace (logs and all) recursively, in-package.
eri_dir_delete("atlantis", azcontainer = con)
```

> **Real logs are not disposable.** The error and data-quality trail across real countries is how the
> team debugs the pipeline and audits what happened. Resolve items; don't delete them. (`atlantis` is
> safe only because it is invented.)

## What's next

You can now find, filter, and close out the error/data-quality backlog — solo or covering for a
teammate. The logs you are reading are written by the workflows in the other guides:

- [Ingesting a surveillance dataset](da-ingest-guide.html) — where DQ flags and approval logs come from.
- [Working with ODK Central](da-odk-guide.html) — sync logs.

See the [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md) and the
**Logs & triage** group in the [reference](../reference/index.html).
