---
title: "Onboarding a new country, disease, or data type (for data analysts)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Onboarding a new country, disease, or data type (for data analysts)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

::: {.eri-guide-meta}
**Walkthrough** &middot; ~20 min &middot; needs: Azure &middot; sandbox-safe: [yes]{.eri-sandbox-yes} (runs on `atlantis`)
:::

> **Just need to scaffold the space, not learn the functions?** Run
> [`eri_do()`](../reference/eri_do.html) in the console and pick "onboard a new country, disease,
> or data type" — it walks you through the dry-run preview and writes the schema template + Azure
> folders for you. It stops there: filling in the schema's disease-specific columns, validating it,
> and submitting it via pull request are still yours to do, and this guide is where you learn that
> part.

Before any data can flow through `erifunctions`, the **space for it has to exist**: a data-quality
schema that describes the data, and the `raw → staged → processed` folders it will move through. This
guide is for a **Data Analyst** standing up that space for a new country, disease, or data type. It is
the **prequel** to the [ingest](da-ingest-guide.html) and [ODK](da-odk-guide.html) guides, do this
once, and those workflows have somewhere to land.

We will practise on a make-believe country, **Atlantis**, so you can run every step and then delete
it. Each `eri_onboard_*` function also has a **dry-run** mode, so you can always look before you touch
anything.

> **You can run this for real.** The commands work against the live Azure system; the final
> [**Clean up**](#clean-up) removes everything you created. Real onboarding follows the exact same
> steps with a real country name.

## The golden rule

> **Onboarding scaffolds; it doesn't finish for you.** `erifunctions` writes you a **schema template**
> with the standard structure and `TODO` markers, and creates the Azure folders. *You* fill in the
> disease-specific details, **validate** the schema, and submit it to the package. Nothing is
> guessed, the template is a correct starting point you complete.

```{=html}
<div class="mermaid">
flowchart TD
    A["Onboard: scaffold schema + folders"] --> B["Fill in the TODO columns"]
    B --> C["eri_schema_validate()"]
    C --> D["Space is ready"]
    D --> E["Ingest data / sync ODK"]
    E --> F["DQ -> stage -> approve"]
    D --> Z["Clean up the sandbox"]
</div>
```

## Before you start {#before-you-start}

1. **R and RStudio** installed.
2. **The package**: `remotes::install_github("thecartercenter/erifunctions")`.
3. **Azure access**: zero-config browser sign-in; see the
   [connections guide](connections-guide.html) if this is your first time.

```{r library}
library(erifunctions)
```

## 1. Look before you leap: the dry run {#dry-run}

Every onboarding command takes `dry_run = TRUE`, which **prints exactly what it would do** and changes
nothing. Always start here:

```{r dry-run}
eri_onboard_country("atlantis", "Atlantis", "malaria", dry_run = TRUE)
#> ℹ Dry run -- nothing will be written or created.
#>   Would write: atlantis_malaria_surveillance_aggregate.yaml
#>   Would create Azure directories:
#>     atlantis/malaria/surveillance/raw
#>     atlantis/malaria/surveillance/staged
#>     atlantis/malaria/surveillance/processed
```

That is the whole footprint: one local schema file, and the three-layer folders in Azure. No
surprises.

## 2. Onboard a surveillance country/disease {#surveillance}

`eri_onboard_country()` is for **surveillance** data (case-level or weekly reporting). The arguments
are the **country code**, the **full country name**, and the **disease**:

```{r onboard-country}
eri_onboard_country("atlantis", "Atlantis", "malaria")
#> ✔ Schema template written to atlantis_malaria_surveillance_aggregate.yaml.
#> ✔ Created 3 Azure directories.
#> ℹ Next steps:
#>   1. Open atlantis_malaria_surveillance_aggregate.yaml and fill in the TODO sections.
#>   2. Run eri_schema_validate('atlantis_malaria_surveillance_aggregate.yaml') to check your edits.
#>   3. Submit the schema via a pull request to add it to the package:
#>      inst/schemas/atlantis_malaria_surveillance_aggregate.yaml
#>   4. Register any ODK forms with eri_odk_register().
#>   5. Pin the package version in your project: renv::snapshot().
```

It did two things: wrote a local schema template, and created `atlantis/malaria/surveillance/{raw,
staged,processed}/` in the data blob.

### Fill in the template

Open `atlantis_malaria_surveillance_aggregate.yaml`. It is a **correct, structured starting point**, the standard
sections are there, with `TODO` markers where your data is unique. Here it is **trimmed for
readability** (your actual file also lists `aliases:` for each column and a few commented options like
`time_grain` and `admin1_name_field`):

```yaml
country: atlantis
disease: malaria
data_source: surveillance
data_type: aggregate
language: en

admin:
  admin1_col: Admin1
  admin2_col: Admin2
  admin1_spatial: spatial/atlantis/atlantis_admin1.shp
  admin2_spatial: spatial/atlantis/atlantis_admin2.shp

temporal:
  year_col: Year
  period_col: EpiWeek
  max_gap: 2

preprocessing:
  - remove_smart_quotes
  - drop_rows_missing_year

columns:
  Year:
    required: true
    type: numeric
    range: [2000, 2035]
  EpiWeek:
    required: true
    type: numeric
    range: [1, 53]
  Admin1:
    required: true
    type: categorical
    # TODO: add allowed_values (list of valid admin1 names)
  Admin2:
    required: true
    type: character

  # TODO: add your disease-specific columns here
  # CasesConfirmed:
  #   required: false
  #   type: numeric
  #   range: [0, 1000000]

consistency:
  # TODO: add cross-field rules here
```

The schema structure, `columns` with `type`/`required`/`range`/`allowed_values`, the `temporal`
block, `consistency` rules, is exactly what the data-quality engine reads. The
[ingest guide](da-ingest-guide.html) shows what each rule *does* when data runs through it.

### Validate your edits

`eri_schema_validate()` checks the **structure**, required sections, valid column types, and that
temporal/consistency rules reference columns that actually exist. Run it on the file you just edited:

```{r validate-good}
eri_schema_validate("atlantis_malaria_surveillance_aggregate.yaml")
#> ✔ Schema 'atlantis_malaria_surveillance_aggregate.yaml' is valid.
#> # A tibble: 0 × 3
#> # ℹ 3 variables: issue_type <chr>, field <chr>, message <chr>
```

An **empty tibble means it's valid**. When something is off, you get a row per problem. Say you fat-
finger two of the column types to `numbr`:

```{r validate-bad}
eri_schema_validate("atlantis_malaria_surveillance_aggregate.yaml")
#> ⚠ 2 issues found in 'atlantis_malaria_surveillance_aggregate.yaml':
#> ✖ Column 'Year' has invalid type 'numbr'. Valid: numeric, character, categorical, date.
#> ✖ Column 'EpiWeek' has invalid type 'numbr'. Valid: numeric, character, categorical, date.

# The returned tibble pinpoints each issue:
#> # A tibble: 2 × 3
#>   issue_type    field                message
#>   <chr>         <chr>                <chr>
#> 1 invalid_value columns.Year.type    Column 'Year' has invalid type 'numbr'. Valid: numeric, …
#> 2 invalid_value columns.EpiWeek.type Column 'EpiWeek' has invalid type 'numbr'. Valid: numeric, …
```

Fix the schema and re-run until you get the clean ✔. That is the green light to use it.

## 3. Onboard a CMR country {#cmr}

**Case Management Reports (CMR)** are the monthly Excel returns from country programs. Onboarding one
writes a CMR schema template and (when you pass `create_dirs = TRUE`) the staging folders under the
combined **`rblf`** disease code (RB + LF, the programmes these reports cover), where
[`eri_stage_cmr()`](da-cmr-guide.html) lands the raw Excel. Dry-run it first:

> **`rblf`/`cmr` is the staging archive, not the canonical home.** Under the
> [source ≠ measure model (ADR-0012)](https://github.com/thecartercenter/erifunctions/blob/main/docs/adr/0012-source-measure-data-model.md),
> CMR is a *format* of the `programmatic` channel. [`eri_split_cmr()`](da-cmr-guide.html) reads the
> staged Excel and routes **each sheet to its own disease and measure**,
> e.g. RB Treatment → `{country}/oncho/programmatic/treatment/`, then you approve **each disease/measure**.
> So `rblf/cmr/` is the transitional landing/archive; the canonical, approved data is per-disease
> (shared sheets like Training route together under the combined `rblf` disease). See the
> [CMR guide](da-cmr-guide.html) for the full upload → stage → split → approve flow.

```{r cmr-dry}
eri_onboard_cmr("atlantis", "Atlantis", create_dirs = TRUE, dry_run = TRUE)
#> ℹ Dry run -- nothing will be written or created.
#>   Would write: atlantis_cmr_schema.yaml
#>     Would create: atlantis/rblf/cmr/raw/
#>     Would create: atlantis/rblf/cmr/staged/
#>     Would create: atlantis/rblf/cmr/processed/
```

Then for real:

```{r cmr-real}
eri_onboard_cmr("atlantis", "Atlantis", create_dirs = TRUE)
#> ✔ CMR schema template written to atlantis_cmr_schema.yaml.
#> ✔ CMR Azure directories created under atlantis/rblf/cmr/.
#> ℹ Next steps:
#>   1. Open atlantis_cmr_schema.yaml and uncomment the sheets your country uses.
#>   2. Match field_code_prefix to the #tag_ row in your CMR Excel file.
#>   3. Submit via pull request to: inst/schemas/cmr/atlantis.yaml
#>   4. Test ingestion with eri_ingest_cmr('your_file.xlsx', country = 'atlantis').
```

A CMR template ships with the common report sheets commented out, uncomment the ones your country
files, and make each sheet's `field_code_prefix` match the `#tag_` row in the Excel file. Then test
with `eri_ingest_cmr()`.

## 4. Onboard an NTD disease (MDA + prevalence) {#ntd}

For an **NTD treatment/survey program**, `eri_onboard_disease()` scaffolds the two schemas that
pattern needs, **MDA** (mass drug administration) and **prevalence**. Note the argument order here is
**disease first, then country**:

```{r onboard-disease}
eri_onboard_disease("schisto", "atlantis")
#> ✔ Schema skeleton written to atlantis_schisto_programmatic_treatment.yaml.
#> ✔ Schema skeleton written to atlantis_schisto_research_prevalence.yaml.
#> ℹ Next steps:
#>   1. Open each file and fill in the TODO sections.
#>   2. Run eri_schema_validate('<path>') to check your edits.
#>   3. Submit via pull request to inst/schemas/.
```

These are **local-only** templates (no Azure folders), an MDA schema with `year`, `round`,
`target_pop`, `treated`, and a `TODO` for the geographic unit, and a matching prevalence schema. Fill,
validate, and submit them exactly as in §2.

> **Contributing the schema to the package, and disease analytics.** Submitting your finished schema
> via pull request, the schema-validation test, and adding disease-specific analytics functions are
> covered in depth in the [**Adding a new program**](adding-a-program.html) vignette. This guide gets
> the space stood up; that one gets your schema merged.

## 5. The space is ready {#ready}

Once your schema validates and the folders exist, the country/disease/data-type is **live in the
system**. You can now:

- **Ingest a surveillance dataset** → the [ingest guide](da-ingest-guide.html).
- **Pull from ODK Central** → the [ODK guide](da-odk-guide.html) (register a form for it, then sync).
- **Make the schema official** → submit it to `inst/schemas/` via pull request so the whole team's
  `load_dq_schema()` finds it.

## 6. Clean up {#clean-up}

Practice run, remove everything you created. Delete the Azure namespace and the local schema files:

```{r cleanup}
con <- get_azure_storage_connection(storage_name = "data")

# Deletes atlantis/malaria/surveillance/* and atlantis/rblf/cmr/* in one go,
# recursively, the in-package path, so you stay in erifunctions instead of
# dropping to raw AzureStor / Storage Explorer.
eri_dir_delete("atlantis", azcontainer = con)

# Remove the local schema templates.
unlink(c("atlantis_malaria_surveillance_aggregate.yaml", "atlantis_cmr_schema.yaml",
         "atlantis_schisto_programmatic_treatment.yaml", "atlantis_schisto_research_prevalence.yaml"))
```

> **Why deleting was fine here:** *Atlantis* is invented. **A real onboarded country is not torn down
> casually**, once data lands in it, those folders are the home of real surveillance records. There
> is deliberately no one-click "un-onboard" command; removing a real space is a considered, manual act.

## What's next

You have stood up a new space three ways, surveillance, CMR, and an NTD program, and validated the
schemas. The data guides take it from here:

- [Ingesting a surveillance dataset](da-ingest-guide.html)
- [Working with ODK Central](da-odk-guide.html)
- [Adding a new program](adding-a-program.html): contribute your schema to the package.

See the [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md) for
the full set, and the [reference](../reference/index.html) for every function grouped by purpose.

::: {.eri-path-nav}
**Data Analyst path, step 2 of 4.** &larr; [Connecting to the services](connections-guide.html)
&middot; Next: [Ingesting a surveillance dataset](da-ingest-guide.html) &rarr;
:::
