---
title: "Quality-check an extract and give a country feedback"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality-check an extract and give a country feedback}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

::: {.eri-guide-meta}
**Walkthrough** &middot; ~10 min &middot; needs: nothing &middot; sandbox-safe: [no]{.eri-sandbox-no} (the QC steps are offline, but the "Notify the team" aside illustrates a real country's Teams channel)
:::

When a country sends a data extract, quality-checking it is only half the job, the other half is
telling them, **specifically and clearly, what needs fixing**. `run_dq_checks()` does both: it
**auto-corrects** the safe, unambiguous things (and records what it changed) and **flags** the rest for
a human. The flags are your feedback to the country.

## The golden rule

> **Auto-corrections are applied and logged; flags are handed back.** A correction is something the
> schema can fix unambiguously (a known spelling, a translation). A flag is something only the country
> can resolve (an impossible age, an unknown locality). Never "fix" a flag by guessing, send it back.

This is offline, the DQ engine runs on a plain data frame. For where DQ sits in the full pipeline see
the [ingest guide](da-ingest-guide.html); for the engine internals and custom checks, the
[DQ pipeline](dq-pipeline.html).

## The extract {#extract}

In practice you'd read the submitted file (`eri_read()` / the raw layer). Here is a small DR malaria
surveillance extract with a few deliberate problems, two values that the schema can normalise, and a
few that it cannot:

```{r data}
extract <- data.frame(
  Year                   = c(2026, 2026, 2026, 2026, 2026),
  EpiWeek                = c(10, 11, 12, 13, 60),          # 60 is out of range
  Age                    = c(34, 5, 200, 41, 28),          # 200 is impossible
  Sex                    = c("Masculino", "Femenino", "Male", "X", "F"),  # X is invalid
  Province_Residence     = c("SANTIAGO", "Santo Domingo", "Azua", "Distrito Nacional", "Nowhere"),
  Municipality_Residence = c("Santiago", "SD", "Azua", "DN", "X"),
  SampleDate             = c("2026-03-01", "2026-03-08", "2026-03-15", "2026-03-22", "2026-03-29"),
  EpiWeekYear            = c(2026, 2026, 2025, 2026, 2026)  # row 3 mismatches Year
)
```

## Run the checks {#run}

Load the schema for this dataset's four axes (`azcontainer = NULL` uses the bundled schema, no Azure),
then run the checks and print the report:

```{r run}
schema <- load_dq_schema("dr", "malaria", "surveillance", "aggregate", azcontainer = NULL)
res    <- run_dq_checks(extract, schema)
#> ✔ DQ checks complete: 3 corrections, 5 flags for review.

dq_report(res)
#> ── Data Quality Report ─────────────────────────────────────────────
#> Shape: 5 rows x 8 columns
#>
#> ── Automated Corrections (3 total) ──
#> • "Sex": 2 corrections (translation)
#> • "Province_Residence": 1 correction (correction)
#>
#> ── Flags Requiring Review (5 total) ──
#> ! Value not in allowed_values list: 2 rows [Sex, Province_Residence] (e.g. X (row 4); Nowhere (row 5))
#> ! Value outside expected range [0, 120]: 1 row [Age] (e.g. 200 (row 3))
#> ! Value outside expected range [1, 53]: 1 row [EpiWeek] (e.g. 60 (row 5))
#> ! Year extracted from SampleDate does not match EpiWeekYear: 1 row [SampleDate] (e.g. 2026-03-15 (row 3))
#>
#> ℹ See `result$flags` for the full row-level detail.
```

## What was fixed automatically {#corrections}

`res$log` is the audit trail of every auto-correction, what changed, in which row, and why. Share it
so the country knows what was normalised on their behalf (it is not an error on their part, but they
should see it):

```{r log}
res$log
#> # A tibble: 3 × 6
#>     row column             original_value corrected_value rule        action
#>   <int> <chr>              <chr>          <chr>           <chr>       <chr>
#> 1     1 Sex                Masculino      Male            translation corrected
#> 2     2 Sex                Femenino       Female          translation corrected
#> 3     1 Province_Residence SANTIAGO       Santiago        correction  corrected
```

## What the country must fix {#flags}

`res$flags` is the feedback list, the rows the schema could **not** resolve. This is what you send
back:

```{r flags}
res$flags
#> # A tibble: 5 × 4
#>     row column             value      issue
#>   <int> <chr>              <chr>      <chr>
#> 1     5 EpiWeek            60         Value outside expected range [1, 53]
#> 2     3 Age                200        Value outside expected range [0, 120]
#> 3     4 Sex                X          Value not in allowed_values list
#> 4     5 Province_Residence Nowhere    Value not in allowed_values list
#> 5     3 SampleDate         2026-03-15 Year extracted from SampleDate does not match EpiWeekYear
```

Turn it into a self-contained handback file with
[`eri_dq_export()`](../reference/eri_dq_export.html), a self-contained HTML (prints cleanly to PDF
from a browser) or markdown file ready to attach to an email or paste into Teams -- one consistent
format instead of an ad hoc table each time:

```{r export}
eri_dq_export(res$flags, country = "dr")
#> ✔ DQ report (5 flags · 5 open) written to '.../dq-report-dr-2026-07-11.html'.
```

## Notify the team / country {#notify}

[`eri_notify_dq()`](../reference/eri_notify_dq.html) posts a one-glance summary (rows, corrections,
flags, the most-flagged columns) to Teams, handy for a shared channel that tracks each country's
submissions:

```{r notify}
eri_notify_dq(res, country = "dr", disease = "malaria")
#> ℹ Sending Teams message via incoming webhook.
```

It posts a compact, at-a-glance summary to the channel your webhook points at:

```
[DQ Report] DR - MALARIA
Rows processed : 5
Corrections    : 3
Flags          : 5
Top flagged    : Age, EpiWeek, Province_Residence, SampleDate, Sex
```

To target a specific Teams channel instead of the default webhook, pass `team =` / `channel =` with a
token (the Graph-API path), see the [connections guide](connections-guide.html#teams).

## What's next

- The cleaned data (`res$data`) continues through the pipeline, see the
  [ingest guide](da-ingest-guide.html) for stage → **approve**.
- The [DQ pipeline](dq-pipeline.html), the check engine, schema rules, and adding `custom_checks`.
- The [epi anomaly guide](epi-dq-guide.html), spikes, gaps, and cross-field/spatial anomalies beyond
  cell-level validity.
- Persist the flags to the shared backlog with `eri_dq_log()`, the
  [log-triage guide](da-logs-guide.html).
- For a CMR workbook specifically, `eri_dq_export()` also renders the richer, per-sheet flags
  tibble from [`eri_cmr_dq_report()`](../reference/eri_cmr_dq_report.html) -- see the
  [CMR guide](da-cmr-guide.html) and the [DQ review guide](da-dq-review-guide.html) for
  [`eri_dq_review()`](../reference/eri_dq_review.html), the interactive triage wrapper that calls it
  automatically.
```
