---
title: "Branded tables, figures, and decks for outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Branded tables, figures, and decks for outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Once data is approved, a lot of the job is turning it into **outputs** — a table for a memo, a figure
for proceedings, a workbook for a partner, a slide deck for a meeting. `erifunctions` ships a small,
consistent **reporting toolkit** so those come out on-brand without fiddling with styling each time.

This guide is the general toolkit (it runs on any data frame). Specific recurring reports get their own
templates as they're defined; the domain figures — **maps** and **epidemic curves** — live in the
[spatial workflow](spatial-workflow.html) and [epi analytics](epi-analytics.html) guides.

## Before you start {#before-you-start}

```{r library}
library(erifunctions)
```

In practice you'd pull approved data first — `eri_query()` for a roll-up (see the
[ad-hoc guide](da-adhoc-guide.html)) or `eri_read()` for one dataset. Here we use a small frame so every
chunk runs as-is:

```{r data}
dat <- data.frame(
  province   = c("North", "South", "East", "West"),
  cases      = c(420, 180, 260, 95),
  tested     = c(1200, 600, 900, 400)
)
dat$positivity <- round(dat$cases / dat$tested * 100, 1)
```

## A branded table {#table}

[`eri_table()`](../reference/eri_table.html) turns a data frame into a styled
[`flextable`](https://davidgohel.github.io/flextable/) — ERI navy header, banded rows, an optional
title and footnote — ready to drop into an HTML/Word/PowerPoint output:

```{r table}
eri_table(
  dat,
  title          = "Malaria cases by province, 2026",
  footnote       = "Source: sandbox data",
  highlight_cols = list(positivity = "#FFC000")
)
```

It returns a `flextable` object; print it in a report, or hand it to the Word/PowerPoint/HTML helpers.
`highlight_cols` is a **named list** mapping a column to a fill colour (`list(col = "#hex")`) — use it to
shade the column you want the eye to land on.

## An on-brand figure {#figure}

Any `ggplot` becomes on-brand by adding [`eri_brand_ggplot_theme()`](../reference/eri_brand_ggplot_theme.html):

```{r plot}
library(ggplot2)
p <- ggplot(dat, aes(x = province, y = cases)) +
  geom_col() +
  labs(title = "Cases by province", x = NULL, y = "Cases") +
  eri_brand_ggplot_theme()
p
```

For domain figures there are purpose-built helpers — `eri_plot_theme("map" | "epicurve")` and
`eri_color_scheme()` (e.g. the standard `"malaria.incidence"` bins), and the `eri_map_*` family for
choropleths and point maps. Those need boundary data; the [spatial workflow](spatial-workflow.html)
guide covers them end-to-end, and [epi analytics](epi-analytics.html) covers `eri_epidemic_curve()`.

## An Excel workbook {#excel}

For a partner who wants the numbers, [`eri_report_excel()`](../reference/eri_report_excel.html) writes a
styled multi-sheet workbook in one call — each named element of `sheets` becomes a tab:

```{r excel}
eri_report_excel(
  sheets = list(Cases = dat),
  path   = "malaria_summary_2026.xlsx",
  title  = "Malaria summary 2026"
)
#> ✔ Workbook saved to 'malaria_summary_2026.xlsx'
```

Need finer control (multiple sheets, added titles)? Build it up with `eri_wb_create()` →
`eri_wb_add_sheet()` → `eri_wb_save()`.

## A slide deck for proceedings {#pptx}

The `eri_pptx_*` family builds a PowerPoint slide by slide — title, sections, tables, and plots — so a
meeting deck is reproducible code, not manual copy-paste:

```{r pptx}
deck <- eri_pptx_create()
deck <- eri_pptx_add_title(deck, "Malaria Summary 2026", subtitle = "ERI")
deck <- eri_pptx_add_table(deck, dat, title = "Cases by province")
deck <- eri_pptx_add_plot(deck, p, title = "Cases by province")
eri_pptx_save(deck, "malaria_summary_2026.pptx")
#> ✔ Presentation saved to 'malaria_summary_2026.pptx'
```

Each `eri_pptx_add_*` call returns the updated deck, so you pipe or reassign as you go. `eri_pptx_create()`
accepts a `template =` to start from a branded master.

## An HTML report {#html}

[`eri_report_html()`](../reference/eri_report_html.html) assembles a self-contained HTML page from
`sections` (text, tables, and plots), for a quick shareable write-up. For a full parameterised report,
`eri_report_qmd_template()` scaffolds a Quarto document you fill in.

## What's next

- **Maps:** the [spatial workflow](spatial-workflow.html) — choropleths, incidence maps, insets.
- **Curves & indicators:** [epi analytics](epi-analytics.html) — `eri_epidemic_curve()`, incidence,
  disease helpers.
- **Where the data comes from:** the [ad-hoc guide](da-adhoc-guide.html) (`eri_query`) and the
  [ingest guide](da-ingest-guide.html).
- The [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md).
```
