---
title: "Final summaries and reports from an ODK survey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Final summaries and reports from an ODK survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Once a survey has been pulled from ODK, cleaned, and **approved**, the last step is the deliverable: a
summary, a results table, a short report. This guide takes an approved survey dataset to a final
summary using the disease helpers, then packages it for sharing.

It picks up where the [ODK guide](da-odk-guide.html) (sync) and the [ingest guide](da-ingest-guide.html)
(stage → approve) leave off, and uses the [reporting toolkit](da-reporting-guide.html) to package the
result. The summary itself is offline.

## The golden rule

> **Report from the approved (`processed/`) dataset, not a working copy.** The number in your report
> should trace to a specific approved file in the catalog. Pull it with `eri_read()` / `eri_query()`,
> summarise, then package, don't re-summarise an ad-hoc export that never went through the gate.

## The survey data {#data}

In practice you'd read the approved survey: `eri_read("ht/lf/research/processed/…")` or an
`eri_query()` roll-up. Here is a small individual-level **LF TAS** (Transmission Assessment Survey)
extract, one row per child tested, with FTS and RDT antigen results:

```{r data}
tas <- data.frame(
  commune    = c(rep("Saut-d'Eau", 6), rep("Mirebalais", 6)),
  fts_result = c("Negative","Negative","Negative","Negative","Positive","Positive",
                 "Negative","Negative","Negative","Positive","Positive","Positive"),
  rdt_result = c("Negative","Negative","Negative","Negative","Negative","Positive",
                 "Negative","Negative","Negative","Negative","Positive","Positive")
)
```

## Summarise with the disease helper {#summary}

[`eri_lf_tas_summary()`](../reference/eri_lf_tas_summary.html) cross-tabulates the FTS and RDT results
into a tidy per-combination summary, optionally per survey unit:

```{r summary}
summary <- eri_lf_tas_summary(tas, fts_col = "fts_result", rdt_col = "rdt_result",
                              group_col = "commune")
summary
#> # A tibble: 6 × 5
#>   commune    fts_result rdt_result     n   pct
#>   <chr>      <chr>      <chr>      <int> <dbl>
#> 1 Mirebalais Negative   Negative       3  50
#> 2 Mirebalais Positive   Negative       1  16.7
#> 3 Mirebalais Positive   Positive       2  33.3
#> 4 Saut-d'Eau Negative   Negative       4  66.7
#> 5 Saut-d'Eau Positive   Negative       1  16.7
#> 6 Saut-d'Eau Positive   Positive       1  16.7
```

Read it per commune by summing the `Positive` FTS rows: Saut-d'Eau has **2 of 6** FTS-positive children,
Mirebalais **3 of 6**. In a real
TAS you compare that antigen-positive count against the survey's **critical cutoff** to decide
pass/fail, the helper gives you the counts the decision rests on.

Other survey measures have their own helpers: `eri_lf_pooled_prev()` for entomology/xenomonitoring
pooled prevalence, and `eri_oncho_program_levels()` / `eri_oncho_status_map()` for oncho, see
[epi analytics](epi-analytics.html).

## Package it for sharing {#package}

The summary is an ordinary tibble, so the [reporting toolkit](da-reporting-guide.html) takes it the rest
of the way, a branded table for the report:

```{r table}
eri_table(
  summary,
  title    = "LF TAS antigen results by commune",
  footnote = "FTS = filariasis test strip; RDT = rapid diagnostic test."
)
```

…a slide for a results meeting:

```{r pptx}
deck <- eri_pptx_create()
deck <- eri_pptx_add_title(deck, "LF TAS Results", subtitle = "Centre Department, 2026")
deck <- eri_pptx_add_table(deck, summary, title = "Antigen results by commune")
eri_pptx_save(deck, "lf_tas_results.pptx")
#> ✔ Presentation saved to 'lf_tas_results.pptx'
```

…or a self-contained HTML write-up with `eri_report_html()`, or an Excel workbook with
`eri_report_excel()`. For a programme-status **map** by evaluation unit, `eri_lf_status_map()` wraps the
spatial helpers (see the [spatial workflow](spatial-workflow.html)).

## What's next

- The [reporting guide](da-reporting-guide.html), the full table/figure/deck/Excel toolkit.
- [Epi analytics](epi-analytics.html): the disease helpers (LF/oncho) and indicators.
- [Working with ODK Central](da-odk-guide.html): pulling the survey in the first place.
- The [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md).
```
