---
title: "A complete research workflow for epidemiologists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A complete research workflow for epidemiologists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This is a hands-on, **start-to-finish walkthrough** of running a research study with
`erifunctions`. It is written for **epidemiologists** — domain experts who are *not*
software developers. You do not need to memorise anything: work through it once, in order,
and you will have done every step a real study needs.

To keep the focus on the *process* rather than the science, we use **`mtcars`** — the little
car dataset that ships with R. It is public, tiny, and safe to copy anywhere, so you can run
every command on your own laptop without touching any real Carter Center data. The
analysis itself is deliberately trivial (a linear regression of fuel economy). **What
matters here is the data management and the research lifecycle**, not the model.

> **You can run this for real.** Every command below works against the live Azure system. If
> you follow along, the very last section — [**Clean up**](#clean-up) — deletes everything you
> created, so you can practise the whole loop end-to-end and leave no trace.

## What you are about to build

We will run a small study twice. First on an **initial dataset** (automatic-transmission cars
only), then we will simulate **"new data arriving"** by expanding to the full dataset and
re-running — exactly the situation you hit when a fresh extract lands months into a project.

```{=html}
<div class="mermaid">
flowchart TD
    A["Start a project"] --> B["Add data + metadata"]
    B --> C["Source the data"]
    C --> D["Analyse (your code)"]
    D --> E["Save outputs + figures"]
    E --> F["Tag a version"]
    F -.->|weeks later| G["Resume + check status"]
    G --> H["New data arrives"]
    H --> D
    F --> Z["Clean up"]
</div>
```

The golden rule that shapes everything:

> **`erifunctions` sources your data reproducibly and keeps it disciplined. _You_ do the
> analysis.** The package gets data in and out, tracks where every input came from, and freezes
> citable versions. The modelling, the epidemiology, the judgement calls — those live in *your*
> project's analysis code, not in the package.

## Before you start {#before-you-start}

You need four things, once:

1. **R and RStudio** installed on your laptop.
2. **The package**, installed from the Carter Center repository:
   ```{r install}
   install.packages("remotes")
   remotes::install_github("thecartercenter/erifunctions")
   ```
3. **Azure access.** There is nothing to configure. The first time a command needs Azure, your
   browser opens and you sign in with your Carter Center account; the sign-in is remembered for
   the rest of the session. (No keys, no environment variables — see `?get_azure_storage_connection`.)
4. **A GitHub account**, and [GitHub Desktop](https://desktop.github.com/) if you prefer clicking
   to typing. We will use it below.

One optional comfort setting: `erifunctions` narrates what it is doing. If you find it chatty,
turn it down for the session (or permanently in your `.Rprofile`):

```{r verbosity}
library(erifunctions)

eri_verbosity("quiet")   # headline results, warnings and errors only
# eri_verbosity("full")  # the default: step-by-step confirmations
```

## 1. Start a project {#start-a-project}

A study is its own self-contained project folder *and* its own GitHub repository (this keeps
each study independently reproducible — see ADR-0006). One command builds the whole skeleton:

```{r scaffold}
eri_research_scaffold(
  name        = "mtcars_study",
  country     = "us",
  disease     = "demo",
  description  = "A worked example: linear regression of fuel economy (mpg)."
)
#> ✔ Scaffolded research project "mtcars_study" at ./mtcars_study.
#>   Next: cd mtcars_study, then git init and renv::init() (see README).
```

This creates a `mtcars_study/` folder containing:

```
mtcars_study/
├── analysis/
│   └── workflow.qmd        # ← write your analysis here
├── data/                   # inputs you pull in (never committed to git)
├── figs/                   # figures
├── outputs/                # model results, tables
├── research.yaml           # the project's logbook + provenance record
├── README.md
├── .gitignore              # already set up to keep data/ out of git
└── .github/workflows/
    └── ci.yaml             # the "reproducibility check" (see §2)
```

Now open the project in RStudio: **File ▸ Open Project**, navigate into `mtcars_study/`, and
open it. Keeping RStudio rooted at the project (rather than opening loose files) matters for the
next step.

## 2. Make it a GitHub repository with the reproducibility check {#github-repo}

Your study should live on GitHub so it is backed up, shareable, and — crucially — so the
**reproducibility check** runs. That check (`.github/workflows/ci.yaml`, created for you above)
confirms that anyone can rebuild your exact software environment and load `erifunctions`. It
stays dormant until you give it a `renv.lock` to check against.

### Put the project under version control

You can do all of this **from inside R** — no command line needed:

```{r git}
install.packages("usethis")
usethis::use_git()      # turns the folder into a git repo and makes the first commit
usethis::use_github()   # creates the GitHub repo and pushes (asks to authorise once)
```

> **Prefer clicking? Use GitHub Desktop.** Open GitHub Desktop ▸ **File ▸ Add Local
> Repository**, choose your `mtcars_study/` folder, and click **"create a repository"** when it
> offers. Then **Publish repository** to put it on GitHub. From then on you **commit** and
> **push** with the buttons in GitHub Desktop instead of typing git commands. Everywhere below
> where we say "commit", that is the GitHub Desktop **Commit to main** button.

### Pin your software versions (this turns the check on)

`renv` records the exact version of every package the study uses, so the analysis runs the same
way next year as it does today:

```{r renv}
install.packages("renv")
renv::init()                                   # start tracking package versions
renv::install("thecartercenter/erifunctions")  # install the package INTO the project
renv::snapshot(type = "all")                    # write renv.lock
```

Then **commit `renv.lock`** (and the `renv/` activation files). The moment that lockfile lands on
GitHub, the reproducibility check goes green on every push — it restores your pinned environment
on a clean machine and confirms `erifunctions` loads.

> **Two `renv` gotchas worth avoiding.**
>
> - Install **`thecartercenter/erifunctions`**, not a personal fork, *before* you snapshot —
>   otherwise the lockfile pins the wrong source and a collaborator restoring it gets the wrong
>   package.
> - Use `renv::snapshot(type = "all")` so packages you only use for analysis (e.g. `ggplot2`) are
>   captured too, not just the ones the package itself imports.

## 3. Add your data and give it metadata {#add-data}

Data does not live in your git repository — it lives in Azure, with a record of what it is. We
register a file as an **artifact**: the file plus its *metadata* (a name, a type, a description,
a version). This is how teammates later discover and trust it.

First, create our **initial dataset** — automatic-transmission cars only (`am == 0`) — and save
it locally:

```{r make-v1, eval=TRUE}
cars_v1 <- mtcars[mtcars$am == 0, ]
nrow(cars_v1)   # 19 cars to start with
```

```{r write-v1}
write.csv(cars_v1, "data/cars.csv", row.names = TRUE)
```

Now upload it as an artifact with its metadata:

```{r artifact-upload}
eri_artifact_upload(
  local_path  = "data/cars.csv",
  name        = "mtcars_study_data",
  type        = "study_data",
  description = "mtcars fuel-economy study data — automatic transmissions only.",
  version     = "1.0"
)
#> ✔ Artifact "mtcars_study_data" (study_data) uploaded to artifacts/study_data/mtcars_study_data/cars.csv.
```

That `name`, `type`, `description`, and `version` **are the metadata** — they travel with the
file and show up in `eri_artifact_list()` so anyone can see what it is without opening it.

> **"Metadata" and "tags" are two different things — don't mix them up.**
>
> | | What it describes | When you set it | Function |
> |---|---|---|---|
> | **Artifact metadata** | *a single file* — what it is, its version | every time you upload data | `eri_artifact_upload()` |
> | **A version tag** | *the whole study* — data **+** analysis code **+** outputs, frozen together and citable | at milestones only (see §10) | `eri_research_tag()` |
>
> Think of metadata as the **label on a jar**, and a version tag as a **photograph of the entire
> kitchen** at one moment. We create our first tag in §10.

## 4. Source the data into your project {#source-data}

Uploading put the data in Azure; now pull a copy *into your project* to work on. We use
`eri_research_pull()` and hand it the artifact's path. Pulling this way (rather than just reading
a local file) records **where the data came from and when** in `research.yaml`, and sets up the
**reversible updates** you'll rely on in §11 when fresh data arrives:

```{r research-pull}
eri_research_pull(path = "artifacts/study_data/mtcars_study_data/cars.csv", dest = "data")
#> ✔ Pulled 1 file from artifacts/study_data/mtcars_study_data/cars.csv to data.
```

> **Two ways to pull, on purpose.** `eri_artifact_pull("mtcars_study_data")` is the quick
> registry shortcut. Here we pull by *path* with `eri_research_pull()` because it also records the
> pull in `research.yaml`, archives any superseded version, and counts updates — exactly the
> lifecycle this guide demonstrates in §11.

> **Never commit `data/` to git.** The scaffold's `.gitignore` already excludes it. Data lives in
> Azure; your repository holds only *code* and the `research.yaml` record of what was pulled. For
> `mtcars` it would not matter, but this is the habit that keeps real, protected country data safe
> — see [Clean up](#clean-up).

## 5. Where to run your analysis {#run-analysis}

Your analysis goes in **`analysis/workflow.qmd`** in the project — that is what the scaffold
seeded it for, and what gets version-controlled and reproducibility-checked. Open it and write
ordinary R. Read the data you just pulled, and fit the model:

```{r analyse}
cars <- read.csv("data/cars.csv", row.names = 1)

model <- lm(mpg ~ wt + hp + disp, data = cars)
summary(model)
```

```{r analyse-output, eval=TRUE, echo=FALSE}
# Rendered output for the site (the data/ file above doesn't exist at build time).
cars  <- mtcars[mtcars$am == 0, ]
model <- lm(mpg ~ wt + hp + disp, data = cars)
summary(model)
```

That is the whole analysis. Notice that **none of it used `erifunctions`** — the modelling is
yours. The package's job starts again when you want to *save* and *freeze* what you found.

## 6. Save your analysis and figures {#save-outputs}

Push your results back to Azure so they are safe and tied to the project. Save the fitted model
object, and a figure:

```{r make-fig}
library(ggplot2)
p <- ggplot(cars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Fuel economy vs. weight (automatic cars)",
       x = "Weight (1000 lbs)", y = "Miles per gallon")
ggsave("figs/mpg_vs_weight.png", p, width = 6, height = 4, dpi = 150)
```

```{r save-outputs}
eri_research_upload_output(model, "mtcars_lm.qs2")
#> ✔ Output "mtcars_lm.qs2" uploaded to research/mtcars_study/outputs/mtcars_lm.qs2.

eri_research_upload_figure(
  "figs/mpg_vs_weight.png",
  caption = "MPG vs. weight, automatic cars (initial dataset)."
)
#> ✔ Figure "mpg_vs_weight.png" uploaded to research/mtcars_study/outputs/figs/mpg_vs_weight.png.
```

As you make decisions, jot them in the project's **logbook**. These notes are timestamped and
stay with the project — they are how "why did we drop those rows?" gets answered a year later:

```{r log}
eri_research_log("Started with automatic transmissions only (am == 0); 19 cars.")
eri_research_log("Fit lm(mpg ~ wt + hp + disp). Weight is the dominant predictor.")
```

## 7. Look at the state of your project {#check-state}

At any point, ask the project where it stands. `eri_research_status()` prints what data you have
pulled, how many times each input has been updated, and how many outputs, snapshots, and tags
exist:

```{r status}
eri_research_status()
#> ── Research project: mtcars_study (us/demo) ──
#> • 1 tracked input (1 pull, 0 artifacts)
#> • 2 outputs, 0 snapshots, 0 tags, 0 promotions
#> # A tibble: 1 × 5
#>   kind  source                                          pulled_at            updates archived
#>   <chr> <chr>                                           <chr>                  <int> <lgl>
#> 1 pull  artifacts/study_data/mtcars_study_data/cars.csv 2026-06-17T15:01:00Z       0 FALSE
```

`eri_research_resume()` is the shorter "where was I?" view — last pull, last logbook entry,
snapshot count. Use it at the **top of every work session** (more on that in §9).

## 8. Save a figure again later (updating it) {#update-figure}

Figures evolve. When you improve one, just **re-upload it under the same filename** — it replaces
the previous version in Azure and the project keeps pointing at the current figure. There is no
special "update" command; uploading the same name *is* the update. (We will do exactly this in
§11 when the data changes.)

## 9. Pause work — and pick it up weeks later {#pause-and-resume}

Real studies stop and start. Here is how to put one down safely and find your way back.

### Pausing

Before you walk away, take a **snapshot** — a timestamped freeze of your whole `data/` directory
in Azure, so the exact inputs are recoverable even if a file changes later. Leave yourself a note,
commit your code, and you are done:

```{r snapshot}
eri_research_snapshot(label = "before-break")
#> ✔ Snapshot of 1 file saved to research/mtcars_study/snapshots/20260617T150500Z/.

eri_research_log("Pausing here. Initial model done; awaiting the fuller dataset.")
```

Then **commit** your code (the `usethis::use_git()` first commit, or the **Commit to main** button
in GitHub Desktop) and close your laptop. Nothing is lost.

### Resuming, weeks later

Reopen the **project** in RStudio (open the project, not loose files, so `renv` re-activates your
pinned environment automatically). Then ask the project to reintroduce itself:

```{r resume}
library(erifunctions)
eri_research_resume()
#> ✔ Project: 'mtcars_study' (us / demo)
#>   Azure:      research/mtcars_study/
#>   Last pull:  2026-06-17T15:00:00Z
#>   Last log:   [2026-06-17T15:05:00Z] Pausing here. Initial model done; awaiting the fuller dataset.
#>   Snapshots:  1

eri_research_status()   # the fuller picture: inputs, outputs, tags
```

Between `resume()` (what was I doing?) and `status()` (what do I have?), you are oriented in
under a minute, however long it has been.

## 10. When — and how — to tag a version {#tagging}

A **tag** freezes the whole study at a moment — the data snapshot, the exact analysis code
commit, and the outputs — under a name you can cite. It is the answer to *"reproduce the numbers
in the report dated March."*

> **When should you tag?** At **milestones**, not on a schedule. Good moments: you have finished
> an analysis you will share or present; you are about to make a big change; you are submitting or
> publishing. **Do not tag every day** — a tag marks *significant progress*, and tags are
> immutable, so a flurry of throwaway tags just adds noise. (Snapshots, from §9, are the cheap,
> frequent safety net; tags are the rare, citable landmarks.)

We have just finished our first real result, so tag it. Make sure your analysis code is
**committed first** (the tag records the commit, and warns you if the working tree has
uncommitted changes):

```{r tag-v1}
eri_research_tag(
  "mtcars-am0-v1",
  description = "Initial model: mpg ~ wt + hp + disp, automatic transmissions only."
)
#> ✔ Tagged "mtcars-am0-v1": snapshot 20260617T150500Z + code "a1b2c3d".
```

That tag now binds *this data + this code + these outputs* together forever. Anyone with the
label can reconstruct exactly what you did.

## 11. New data arrives — update and re-run {#new-data}

Months in, a fuller dataset lands: we are no longer restricted to automatic cars. This is the
moment most workflows handle badly — overwriting the old data and losing the trail. `erifunctions`
handles it cleanly: the old version is archived, the change is counted, and your earlier tag still
reproduces the original.

Build the **expanded dataset** (drop the `am == 0` filter — all cars now) and re-upload it under
the **same artifact name**, bumping the version. Re-using the name is an *update*, not a duplicate:

```{r make-v2, eval=TRUE}
cars_v2 <- mtcars          # the full dataset now — both transmission types
nrow(cars_v2)              # 32 cars, up from 19
```

```{r upload-v2}
write.csv(cars_v2, "data/cars.csv", row.names = TRUE)

eri_artifact_upload(
  local_path  = "data/cars.csv",
  name        = "mtcars_study_data",      # same name → updates the existing artifact
  type        = "study_data",
  description = "mtcars fuel-economy study data — all transmissions (expanded).",
  version     = "2.0"
)
```

Now pull the new version into the project. Because you have pulled this input before,
`erifunctions` **archives the previous copy**, swaps in the new one, and records that this input
has now been updated:

```{r pull-v2}
eri_research_pull(path = "artifacts/study_data/mtcars_study_data/cars.csv", dest = "data")
#> ℹ Archived 1 prior version to data/_archive/20260701T090000Z.
#> ✔ Pulled 1 file from artifacts/study_data/mtcars_study_data/cars.csv to data.

eri_research_status()
#> ── Research project: mtcars_study (us/demo) ──
#> # A tibble: 1 × 5
#>   kind  source                                          pulled_at            updates archived
#>   <chr> <chr>                                           <chr>                  <int> <lgl>
#> 1 pull  artifacts/study_data/mtcars_study_data/cars.csv 2026-07-01T09:00:00Z       1 TRUE
```

See how `updates` ticked to `1` and `archived` is now `TRUE` — the original 19-car extract is
safely archived, not gone. Re-run the analysis on the fuller data, update the figure (same
filename — §8), log the change, and **tag the new version**:

```{r rerun, eval=TRUE}
cars <- mtcars
model <- lm(mpg ~ wt + hp + disp, data = cars)
coef(model)
```

```{r rerun-save}
library(ggplot2)
p <- ggplot(cars, aes(wt, mpg)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Fuel economy vs. weight (all cars)")
ggsave("figs/mpg_vs_weight.png", p, width = 6, height = 4, dpi = 150)  # same name = update

eri_research_upload_output(model, "mtcars_lm.qs2")
eri_research_upload_figure("figs/mpg_vs_weight.png",
                           caption = "MPG vs. weight, all cars (expanded dataset).")
eri_research_log("Expanded to the full dataset (32 cars). Re-fit and updated the figure.")

# Snapshot the new data first, so the tag binds *this* (full) dataset rather than the
# earlier "before-break" snapshot.
eri_research_snapshot(label = "full-data")
eri_research_tag(
  "mtcars-full-v1",
  description = "Re-run on the full dataset (all transmissions)."
)
```

You now have **two citable versions**: `mtcars-am0-v1` (the original) and `mtcars-full-v1` (the
update). The first still reproduces the original analysis exactly, even though the data has moved
on — which is the whole point.

## 12. Clean up {#clean-up}

If you ran this walkthrough for practice, remove everything it created so you leave no trace in
Azure. This deletes the project folder, all snapshots and tags, and the artifact:

```{r cleanup}
con <- get_azure_storage_connection(storage_name = "data")

# eri_dir_delete() removes a non-empty directory recursively in one call — the
# in-package path, so cleanup stays in erifunctions instead of raw AzureStor.
eri_dir_delete("research/mtcars_study", azcontainer = con)
eri_dir_delete("artifacts/study_data/mtcars_study_data", azcontainer = con)
```

Confirm nothing of yours remains, then delete the local folder:

```{r cleanup-verify}
eri_research_list()       # mtcars_study should be gone
eri_artifact_list()       # mtcars_study_data should be gone

unlink("../mtcars_study", recursive = TRUE)   # remove the local project folder
```

> **Why we could delete freely here:** `mtcars` is public sample data. **Real Carter Center data
> is different** — country surveillance and study data must never be deleted casually, copied off
> Azure, or committed to git. The discipline this guide builds (data stays in Azure, pulled with
> provenance, frozen with tags) is exactly what keeps protected data safe. When in doubt, ask
> before you delete.

## What's next

You have now done the full research lifecycle: start, version-control, add and source data,
analyse, save, tag, pause, resume, handle new data, and tidy up.

This is the first of a planned set of short, task-oriented guides — **one for each common
job** an analyst or epidemiologist does (running a monthly country report, ingesting a
surveillance dataset, onboarding a new country, and more). See
[the guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md)
for what exists today and what is coming.

For the full menu of functions grouped by purpose, see the
[reference](../reference/index.html).
