---
title: "Reconciling free-text localities to admin units (for epidemiologists)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reconciling free-text localities to admin units (for epidemiologists)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Surveillance data arrives with **messy, free-text place names** — `"Jínova"`, `"JUAN DE HERRERA"`,
`"S. Juan"` — that have to be matched to the **canonical admin units** in an authoritative boundary
before you can aggregate cases, compute incidence, or map anything. Doing that by hand is slow and
error-prone. This guide, for **epidemiologists**, shows how
[`eri_spatial_reconcile()`](../reference/eri_spatial_reconcile.html) does it: match what it safely can,
geocode the rest, and **flag anything it isn't sure about** so you check it before trusting it.

> **This is sourcing, not analysis.** Reconciliation is the data-preparation step many studies do by
> hand. The matching, windowing, and modelling stay in *your* research project (see the
> [research guide](epi-research-guide.html)); this just gets clean, canonical localities in.

```{=html}
<div class="mermaid">
flowchart TD
    A["Messy free-text localities"] --> B["Pass 1: string match (offline)"]
    B -->|matched| M["✓ matched"]
    B -->|no match| C["Pass 2: geocode the residual"]
    C --> D["Point-in-polygon + trust guard"]
    D -->|parents agree| G["✓ geocoded"]
    D -->|parents disagree / low confidence| R["⚠ geocoded_review"]
    C -->|no result| U["✗ unresolved"]
</div>
```

## Before you start {#before-you-start}

- `remotes::install_github("thecartercenter/erifunctions")`, plus the **`sf`** package
  (`install.packages("sf")`).
- The **string-match** step runs fully offline. The optional **geocoding** step uses **OpenStreetMap
  (Nominatim)** by default — keyless, so no setup — though it makes network calls. (Higher-quality
  geocoding via Google is available with a key; see [the end](#real-data).)

```{r library}
library(erifunctions)
library(sf)
```

## 1. The reference boundary and your messy data {#inputs}

Reconciliation needs an **authoritative boundary** — an `sf` object whose columns hold the canonical
admin names. In real work you load it with
[`eri_spatial_load()`](../reference/eri_spatial_load.html); for this guide we build a tiny one by hand
(two real places so the live geocoder has somewhere to land):

```{r boundary}
box <- function(xmin, xmax, ymin, ymax) {
  st_polygon(list(matrix(c(xmin,ymin, xmax,ymin, xmax,ymax, xmin,ymax, xmin,ymin),
                         ncol = 2, byrow = TRUE)))
}
boundary <- st_sf(
  adm3_name = c("Boston", "Cambridge"),
  adm2_name = c("Suffolk", "Middlesex"),     # the authoritative county for each
  adm1_name = c("Massachusetts", "Massachusetts"),
  geometry  = st_sfc(box(-71.10, -70.99, 42.230, 42.350),
                     box(-71.16, -71.06, 42.350, 42.400), crs = 4326)
)
```

And the incoming data — five localities with the usual problems (case, a typo, names not in the
boundary, and one that is pure noise):

```{r messy-data, eval=TRUE}
messy <- tibble::tibble(
  locality = c("boston", "Cambrige", "Fenway Park", "MIT", "Zzqxborough Nowhere"),
  county   = c("Suffolk", "Middlesex", "Suffolk",   "Suffolk", "Suffolk"),
  state    = rep("Massachusetts", 5),
  cases    = c(12L, 7L, 4L, 3L, 1L)
)
messy
```

You describe the hierarchy with two **parallel** vectors, **finest → coarsest**: your data's columns
(`loc_cols`) and the matching boundary columns (`admin_cols`).

## 2. Pass 1 — string match (offline) {#match}

First, match what you can with no network at all. Setting `method = NULL` disables geocoding entirely,
and `max_dist = 1` allows a one-character typo at the finest level:

```{r match}
matched <- eri_spatial_reconcile(
  messy,
  loc_cols   = c("locality", "county", "state"),
  admin_cols = c("adm3_name", "adm2_name", "adm1_name"),
  shapefile  = boundary,
  method     = NULL,    # string match only — fully offline
  max_dist   = 1L       # tolerate a 1-character typo in the finest name
)
#> ✔ Reconciled 5 distinct localities: 2 matched, 0 geocoded, 0 need review, 3 unresolved.

matched[, c("locality", "county", "state", "reconcile_status")]
#> # A tibble: 5 × 4
#>   locality            county    state         reconcile_status
#>   <chr>               <chr>     <chr>         <chr>
#> 1 Boston              Suffolk   Massachusetts matched
#> 2 Cambridge           Middlesex Massachusetts matched
#> 3 Fenway Park         Suffolk   Massachusetts unresolved
#> 4 MIT                 Suffolk   Massachusetts unresolved
#> 5 Zzqxborough Nowhere Suffolk   Massachusetts unresolved
```

Two wins already, offline: `"boston"` matched `Boston` (case-insensitive, accent-insensitive), and the
typo `"Cambrige"` matched `Cambridge` (within `max_dist`). For a `matched` row the **whole hierarchy**
is rewritten to the boundary's canonical spelling — not just the locality, but the county and state too
— so a misspelled parent gets fixed as well. The other three don't appear in the boundary, so they're
`unresolved` — they're the residual to geocode.

## 3. Pass 2 — geocode the residual {#geocode}

Now let geocoding handle the rows that didn't match. `method = "osm"` uses keyless OpenStreetMap;
`country_name` is appended to each address to anchor it:

```{r geocode}
result <- eri_spatial_reconcile(
  messy,
  loc_cols     = c("locality", "county", "state"),
  admin_cols   = c("adm3_name", "adm2_name", "adm1_name"),
  shapefile    = boundary,
  country_name = "USA",
  method       = "osm",
  max_dist     = 1L
)
#> ℹ Geocoding 3 unmatched localities via "osm"...
#> ✔ Reconciled 5 distinct localities: 2 matched, 1 geocoded, 1 need review, 1 unresolved.
#> ! 1 geocode flagged "geocoded_review" (low-confidence or parent mismatch) -- verify before use.

result[, c("locality", "county", "longitude", "latitude", "reconcile_status")]
#> # A tibble: 5 × 5
#>   locality            county    longitude latitude reconcile_status
#>   <chr>               <chr>         <dbl>    <dbl> <chr>
#> 1 Boston              Suffolk         NA      NA   matched
#> 2 Cambridge           Middlesex       NA      NA   matched
#> 3 Boston              Suffolk      -71.1     42.3  geocoded
#> 4 MIT                 Suffolk      -71.1     42.4  geocoded_review
#> 5 Zzqxborough Nowhere Suffolk         NA      NA   unresolved
```

Three different outcomes for the three residual rows:

- **`Fenway Park` → `geocoded`.** It geocoded to a point inside the **Boston/Suffolk** polygon, which
  *agrees* with the county you supplied — so it's trusted: the name is rewritten to the canonical
  `Boston` and you get coordinates.
- **`MIT` → `geocoded_review`.** It geocoded fine, but the point falls in the **Cambridge/Middlesex**
  polygon — while you claimed **Suffolk**. That disagreement is the **trust guard** firing: the
  coordinates are kept for you to inspect, but the names are **left exactly as you supplied them** (not
  silently "corrected"). Maybe MIT really is across the county line from where you expected — or maybe
  the county was mis-entered. *You* decide.
- **`Zzqxborough Nowhere` → `unresolved`.** The geocoder returned nothing. No coordinates, nothing
  changed.

You don't have to take "Cambridge/Middlesex" on faith. `eri_spatial_reconcile()` returns the admin
units the geocoded point actually fell into as `geocoded_<admin_col>` columns — so a flagged row tells
you *exactly* what it disagreed with, with no manual `sf::st_join()` to re-derive it:

```{r inspect-review}
result[result$reconcile_status == "geocoded_review",
       c("locality", "county", "geocoded_adm3_name", "geocoded_adm2_name")]
#> # A tibble: 1 × 4
#>   locality county  geocoded_adm3_name geocoded_adm2_name
#>   <chr>    <chr>   <chr>              <chr>
#> 1 MIT      Suffolk Cambridge          Middlesex
```

Your `county` is still **Suffolk** (untouched — the trust guard never overwrote it), but
`geocoded_adm2_name` shows the point landed in **Middlesex**. That side-by-side is the whole conflict,
ready for you to confirm or fix. (These columns are `NA` for `matched`, `unresolved`, and
outside-polygon rows.)

## 4. What the statuses mean — and what to do {#statuses}

| `reconcile_status` | What happened | What you do |
|---|---|---|
| `matched` | Name matched the boundary (offline) | Trust it — canonical name assigned |
| `geocoded` | Geocoded **and** the admin parents agree with your data | Trust it — canonical name + coordinates assigned |
| `geocoded_review` | Geocoded, but the assigned admin unit **disagrees** with your claim (or the service flagged low confidence) | **Inspect** the `geocoded_*` columns to see what it disagreed with. Coordinates kept; names left untouched. Confirm or fix before using |
| `unresolved` | No match, and either no geocode result **or** a point that fell outside every boundary (such rows may still carry coordinates) | Chase the source, or exclude with a note |

The golden habit: **`matched` and `geocoded` are safe to roll up; `geocoded_review` and `unresolved`
need a human** before they enter an analysis.

## 5. Feed it downstream {#downstream}

The result is a plain tibble — your original rows, with canonical names filled in where confident and
`longitude`/`latitude` for anything geocoded. From here it is ordinary analysis (in *your* research
project, not the package): aggregate the trusted rows, join to case counts, or map them. For mapping
and spatial joins on the canonical units, see the
[spatial workflow](spatial-workflow.html) vignette.

```{r downstream}
# e.g. roll up only the rows you trust
trusted <- result[result$reconcile_status %in% c("matched", "geocoded"), ]
aggregate(cases ~ county, data = trusted, sum)
```

## 6. Using this on real data {#real-data}

- **Boundaries:** replace the hand-built `boundary` with
  `eri_spatial_load("dr", level = 3)` (or your country/level) to reconcile against the real admin
  units.
- **Better geocoding:** OpenStreetMap is free and keyless but coarse. For higher accuracy use
  `method = "google"` — it needs your own API key in `.Renviron`
  (`GOOGLEGEOCODE_API_KEY=...`); `eri_spatial_reconcile()` checks for it up front and tells you exactly
  what to add if it's missing. **Never put a key in a script or commit it.** Google also returns a
  `partial_match` flag, which feeds the same `geocoded_review` trust guard.
- **No clean-up needed:** reconciliation is entirely in-memory — it reads your boundary and writes
  nothing to Azure.

## What's next

You can now turn a column of messy place names into canonical admin units with coordinates, with the
uncertain ones flagged rather than hidden. From here:

- [Running a research study](epi-research-guide.html) — where reconciled data feeds the analysis.
- [Spatial workflow](spatial-workflow.html) — mapping and joining on the canonical units.

See the [guide index](https://github.com/thecartercenter/erifunctions/blob/main/docs/guides.md) and the
[reference](../reference/index.html) for the full set.
