---
title: "Epidemiological Analytics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Epidemiological Analytics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette covers the epidemiological helper functions for malaria, lymphatic
filariasis, and onchocerciasis analyses — from computing incidence rates and
epiweek dates to LF pooled prevalence and onchocerciasis programme-status maps.

## Malaria: incidence and epidemic curves

### Incidence rate

```{r incidence}
library(erifunctions)
library(dplyr)

weekly <- result$data |>
  group_by(year, epiweek, province) |>
  summarise(n_cases = n(), .groups = "drop") |>
  left_join(population_data, by = c("year", "province")) |>
  mutate(rate_per_1000 = eri_incidence_rate(n_cases, population, multiplier = 1000))
```

Rows with `population <= 0` or `NA` values automatically return `NA` to avoid
silent division errors.

### Epiweek dates

Convert CDC epiweek numbers to calendar dates for time-series plots:

```{r epiweek}
weekly <- weekly |>
  mutate(week_start = eri_epiweek_date(year, epiweek, week_start = "Sunday"))

library(ggplot2)
ggplot(weekly, aes(week_start, n_cases, colour = province)) +
  geom_line() +
  labs(x = NULL, y = "Cases", title = "Weekly malaria cases by province") +
  eri_brand_ggplot_theme()
```

### Study week

`eri_study_week()` expresses epiweeks relative to an index event — useful for
intervention analyses such as IRS campaigns:

```{r study-week}
irs_date <- as.Date("2023-06-15")

weekly <- weekly |>
  mutate(study_week = eri_study_week(year, epiweek, index_date = irs_date))

# week < 0 = before IRS; week >= 1 = after IRS
pre_post <- weekly |>
  mutate(period = ifelse(study_week < 0, "Pre-IRS", "Post-IRS"))
```

### Epiweek from date

Convert individual case dates to epiweek numbers for aggregation:

```{r date-to-epiweek}
cases <- cases |>
  mutate(epiweek = eri_date_to_epiweek(sample_date))
```

## Branded ggplot theme and colour palette

All ERI figures should use `eri_brand_ggplot_theme()` and colours from
`eri_brand_colors()`:

```{r theme}
cols <- eri_brand_colors()
# navy="#44546A"  blue="#4472C4"  orange="#ED7D31"
# gold="#FFC000"  green="#70AD47" light_blue="#5B9BD5"  gray="#A5A5A5"

ggplot(weekly, aes(week_start, rate_per_1000)) +
  geom_col(fill = cols[["blue"]]) +
  eri_brand_ggplot_theme() +
  labs(title = "Malaria incidence rate", y = "Cases per 1 000")
```

## Lymphatic filariasis: pooled prevalence

LF TAS data is typically collected in pools (FTS cards). Use
`eri_lf_pooled_prev()` to estimate individual-level prevalence from pool
results using the standard formula
`1 - ((1 - npos/npool)^(1/pool_size))`.

```{r lf-prev}
# Scalar estimate for a single survey
eri_lf_pooled_prev(npos = 3, npool = 100, pool_size = 5)

# Per-commune estimates
commune_prev <- tas_data |>
  group_by(commune) |>
  summarise(
    npos      = sum(fts_result == "Positive"),
    npool     = n(),
    pool_size = mean(pool_size, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  mutate(pooled_prev = eri_lf_pooled_prev(npos, npool, pool_size))
```

The function warns (rather than errors) when `npos > npool` and returns `NA`
for those rows, so batch processing is not interrupted.

## Lymphatic filariasis: MDA coverage

Use the `ht_lf_mda` and `dr_lf_mda` schemas to clean coverage data, then
compute programme indicators:

```{r lf-mda}
schema <- load_dq_schema("ht", "lf_mda", azcontainer = NULL)
result <- run_dq_checks(raw_mda, schema)

coverage <- result$data |>
  mutate(
    coverage_pct = treated / target_pop * 100,
    reached_80   = coverage_pct >= 80
  )

dplyr::count(coverage, year, reached_80)
```

## Onchocerciasis: programme status map

Onchocerciasis uses a standardised five-level programme status framework.
`eri_oncho_program_levels()` returns the ordered factor levels used across
all OEPA visualisations, and `eri_oncho_status_map()` produces a
ready-to-insert `ggplot` object:

```{r oncho}
eri_oncho_program_levels()
# [1] "Non-endemic"
# [2] "Under surveillance"
# [3] "MDA ongoing"
# [4] "MDA stopped - under surveillance"
# [5] "Verified free of transmission"

eus <- eri_spatial_load("oepa", level = 2)

status_map <- eri_oncho_status_map(
  shapefile   = eus,
  status_data = programme_status_2024,
  eu_col      = "eu_name",
  status_col  = "programme_status",
  title       = "OEPA Onchocerciasis Programme Status 2024",
  scale_bar   = TRUE,
  north_arrow = TRUE
)
```

## Styled tables

`eri_table()` produces a Carter Center-branded `flextable` that renders
consistently in PowerPoint, HTML, and Excel:

```{r table}
summary_tbl <- weekly |>
  group_by(province) |>
  summarise(
    total_cases = sum(n_cases),
    mean_weekly = round(mean(n_cases), 1),
    peak_rate   = max(rate_per_1000, na.rm = TRUE)
  )

ft <- eri_table(
  summary_tbl,
  title    = "Malaria summary by province",
  footnote = "Source: DIGEPI surveillance data"
)

# Insert into a PowerPoint presentation
eri_pptx_create() |>
  eri_pptx_add_title("Hispaniola Malaria 2024") |>
  eri_pptx_add_table(summary_tbl, title = "Province summary") |>
  eri_pptx_save("outputs/malaria_2024.pptx")
```
