---
title: "SharePoint Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SharePoint Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

::: {.eri-guide-meta}
**Deep-dive** &middot; ~15 min &middot; needs: Azure &middot; sandbox-safe: [no]{.eri-sandbox-no} (connects to the real SharePoint site)
:::

This vignette covers the full SharePoint read/write pattern using the `eri_sharepoint_*`
functions. These functions require the `Microsoft365R` package:

```{r install, eval=FALSE}
install.packages("Microsoft365R")
```

## 1. Connect to a SharePoint site

```{r connect}
library(erifunctions)

site <- eri_sharepoint_connect("https://cartercenter.sharepoint.com/sites/ERI")
```

`eri_sharepoint_connect()` opens a browser window for interactive authentication the first
time. The token is cached by `AzureAuth` so subsequent calls within a session do not
re-prompt.

## 2. List files in a document library folder

```{r list}
# List the root of the default document library
files <- eri_sharepoint_list(site)

# List a specific subfolder
weekly_files <- eri_sharepoint_list(site, "Shared Documents/Weekly Data/Haiti")
print(weekly_files)
# A tibble with columns: name, size, modified, is_folder, path
```

Filter to find the most recent weekly upload:

```{r filter-list}
library(dplyr)

latest <- weekly_files |>
  filter(!is_folder, grepl("\\.xlsx$", name)) |>
  arrange(desc(modified)) |>
  slice(1)

cat("Latest file:", latest$name, "\n")
cat("Modified:   ", format(latest$modified), "\n")
```

## 3. Read a file directly into R

`eri_sharepoint_read()` downloads the file to a temp location and reads it based on
the extension (`.xlsx`, `.csv`, `.parquet`, `.rds`).

```{r read-excel}
# Read an Excel file
df <- eri_sharepoint_read(site, "Shared Documents/Weekly Data/Haiti/ht_weekly_2024.xlsx")
head(df)
```

```{r read-csv}
# Read a CSV file with extra readr arguments
df_csv <- eri_sharepoint_read(
  site,
  "Shared Documents/Weekly Data/DR/dr_cases_2024.csv",
  col_types = readr::cols(.default = "c")
)
```

```{r read-sheet}
# Read a specific Excel sheet
df_sheet <- eri_sharepoint_read(
  site,
  "Shared Documents/Reports/annual_summary.xlsx",
  sheet = "Malaria"
)
```

For an unknown extension the function returns the local temp path so you can read
it manually:

```{r read-unknown}
path <- eri_sharepoint_read(site, "Shared Documents/Data/custom_format.sav")
# path is a character string -- use haven::read_sav(path) or similar
```

## 4. Upload a finished report to SharePoint

```{r upload}
# Push a local Excel report to the team SharePoint
eri_sharepoint_upload(
  local_path  = "outputs/ht_malaria_weekly_summary.xlsx",
  site        = site,
  folder_path = "Shared Documents/Reports/Haiti/2024"
)
```

The destination folder is created automatically if it does not exist. Use
`overwrite = FALSE` to prevent accidentally replacing a file:

```{r upload-guard}
eri_sharepoint_upload(
  local_path  = "outputs/ht_malaria_weekly_summary.xlsx",
  site        = site,
  folder_path = "Shared Documents/Reports/Haiti/2024",
  overwrite   = FALSE   # errors if the file already exists
)
```

The function returns the SharePoint item URL invisibly:

```{r upload-url}
url <- eri_sharepoint_upload("outputs/report.xlsx", site, "Shared Documents/Reports")
cat("Shared at:", url, "\n")
```

## 5. Combined workflow

The most common pattern is: pull data from SharePoint, run DQ checks, generate a
report, push the report back.

```{r full-workflow}
library(erifunctions)
library(dplyr)

# ---- Step 1: connect ----
site <- eri_sharepoint_connect("https://cartercenter.sharepoint.com/sites/ERI")

# ---- Step 2: pull the latest weekly submission ----
files <- eri_sharepoint_list(site, "Shared Documents/Weekly Data/Haiti")
latest_path <- files |>
  filter(!is_folder, grepl("\\.xlsx$", name)) |>
  arrange(desc(modified)) |>
  slice(1) |>
  pull(path)

raw <- eri_sharepoint_read(site, latest_path)

# ---- Step 3: run DQ checks ----
schema <- load_dq_schema("haiti", "malaria", azcontainer = NULL)
result <- run_dq_checks(raw, schema)

# ---- Step 4: generate a summary Excel report ----
tmp_report <- tempfile(fileext = ".xlsx")
eri_report_excel(
  list(
    data   = list(data = result$data),
    flags  = list(data = result$flags)
  ),
  path = tmp_report
)

# ---- Step 5: push back to SharePoint ----
week_label <- format(Sys.Date(), "%Y-W%V")
eri_sharepoint_upload(
  local_path  = tmp_report,
  site        = site,
  folder_path = paste0("Shared Documents/DQ Reports/Haiti/", week_label)
)
```

## Authentication notes

- Tokens are cached in `~/.AzureR/` by `AzureAuth`. If you are prompted every
  session, check that `~/.AzureR/` is writable.
- For scripted / CI use, prefer service-principal auth configured through
  `Microsoft365R::create_sharepoint_site()` with `auth_type = "client_credentials"`.
- All `eri_sharepoint_*` functions guard against `Microsoft365R` not being
  installed and give a clear install instruction if it is absent.
