---
title: "Spatial Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

::: {.eri-guide-meta}
**Deep-dive** &middot; ~15 min &middot; needs: Azure &middot; sandbox-safe: no (illustrates real country boundaries)
:::

The spatial workflow covers three use cases: loading canonical admin boundaries
from Azure, joining GPS coordinates to those boundaries, and producing
programmatic maps. All functions require the `sf` package.

## Loading admin boundaries

Admin boundaries are stored in Azure at
`spatial/{country}/adm{level}.rds` as validated `sf` objects.
Levels follow the standard hierarchy: 0 = country, 1 = department/region,
2 = province/commune, 3 = municipality/locality, 4 = sub-locality.

```{r load}
library(erifunctions)

# Load Haiti commune (adm2) boundaries
communes <- eri_spatial_load("ht", level = 2)

# Load DR province (adm2) boundaries
provinces <- eri_spatial_load("dr", level = 2)
```

Every object loaded from Azure has been validated to have:
- A defined CRS (WGS 84, EPSG 4326)
- No empty geometries
- A canonical name column `adm{level}_name`

## Uploading new boundaries

When adding a new country or updating boundaries, use `eri_spatial_upload()`.
It validates the shapefile before accepting the upload:

```{r upload}
# Prepare your shapefile first:
#   - Set CRS to 4326 if not already set
#   - Rename the admin name column to adm2_name
#   - Remove any empty geometries

eri_spatial_upload(
  local_path = "data/shapefiles/hti_adm2.shp",
  country    = "ht",
  level      = 2
)
```

The function blocks with a descriptive error if any validation check fails,
explaining exactly what to fix.

## Spatial join: GPS coordinates to admin units

Use `eri_spatial_join()` to assign admin unit names to point data with lat/lon
columns. Rows with missing coordinates are dropped with a warning.

```{r join}
# Load the reference boundary
communes <- eri_spatial_load("ht", level = 2)

# Attach commune and department names to case records
case_data <- eri_spatial_join(
  data       = raw_cases,
  lat_col    = "latitude",
  lon_col    = "longitude",
  shapefile  = communes,
  admin_cols = c("adm2_name", "adm1_name")
)
```

The result is a plain tibble, geometry is dropped automatically, so downstream
analysis does not need to handle `sf` objects.

## Bounding box expansion

When composing maps it is often useful to add a buffer around the study area
so that boundary features are fully visible. `eri_bbox_expand()` takes an `sf`
bounding box and expands it by a specified distance in metres:

```{r bbox}
library(sf)

haiti <- eri_spatial_load("ht", level = 0)
bbox  <- sf::st_bbox(haiti) |>
  eri_bbox_expand(X = 30000, Y = 30000)   # 30 km padding on all sides

# Use the expanded bbox as a crop/map extent
```

## Choropleth maps

Combine `eri_spatial_load()` with `eri_brand_ggplot_theme()` and standard
`ggplot2` / `sf` functions to produce branded choropleth maps.

```{r choropleth}
library(ggplot2)
library(sf)

# Aggregate case counts by commune
case_counts <- dplyr::count(case_data, adm2_name, name = "n_cases")

# Join to boundaries
map_data <- dplyr::left_join(communes, case_counts, by = "adm2_name")

ggplot(map_data) +
  geom_sf(aes(fill = n_cases), colour = "white", linewidth = 0.2) +
  scale_fill_gradient(
    low  = "#E8F4FD",
    high = eri_brand_colors()[["navy"]],
    na.value = "grey90",
    name = "Cases"
  ) +
  labs(title = "Malaria cases by commune") +
  eri_brand_ggplot_theme()
```

## LandScan population raster

LandScan global population rasters are stored in Azure under
`spatial/landscan/`. Use `eri_spatial_load_landscan()` (from `R/spatial.R`)
to load the raster for a given year, then extract population for each admin
unit using the `exactextractr` package:

```{r landscan}
library(terra)
library(exactextractr)

pop_raster <- eri_spatial_load_landscan(year = 2023)
communes   <- eri_spatial_load("ht", level = 2)

# Sum population pixels within each commune
communes$population <- exact_extract(pop_raster, communes, "sum")
```

## Onchocerciasis programme-status maps

For OEPA onchocerciasis data, `eri_oncho_status_map()` wraps the full mapping
pipeline, it joins status data to a shapefile, applies the standard ERI
programme-status colour palette, and adds a scale bar and north arrow:

```{r oncho-map}
eus     <- eri_spatial_load("oepa", level = 2)  # evaluation unit boundaries
status  <- read_dq_result("outputs/oepa_status_2024.parquet")

map <- eri_oncho_status_map(
  shapefile   = eus,
  status_data = status,
  eu_col      = "eu_name",
  status_col  = "programme_status",
  title       = "OEPA Onchocerciasis Programme Status 2024"
)

eri_pptx_create() |>
  eri_pptx_add_plot(map, title = "Programme Status") |>
  eri_pptx_save("outputs/oepa_status_report.pptx")
```
