---
title: "What are you trying to do?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What are you trying to do?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

::: {.eri-guide-meta}
**Desk reference** &middot; ~2 min to scan &middot; needs: nothing &middot; sandbox-safe: [yes]{.eri-sandbox-yes} (reads a bundled file, nothing runs)
:::

Every task below is a real one, generated from the same registry (`inst/registry/task_map.yaml`) that
backs `eri_task_map()`. Find your task, then follow its **Run** call or its **Guide**. New to the
package? [Getting started](getting-started.html) is a better front door; this page is for when you
already know what you're trying to do and just need the entry point.

```{r task-table, eval=TRUE, echo=FALSE, message=FALSE}
tm <- yaml::read_yaml(system.file("registry", "task_map.yaml", package = "erifunctions"))

rows <- do.call(rbind, lapply(tm, function(branch) {
  do.call(rbind, lapply(branch$children, function(leaf) {
    data.frame(
      Category = branch$title,
      Task     = leaf$title,
      Run      = sprintf("`%s`", leaf$call),
      Guide    = if (is.null(leaf$guide)) {
        ""
      } else {
        sprintf("[guide](%s.html)", leaf$guide)
      },
      stringsAsFactors = FALSE
    )
  }))
}))

knitr::kable(rows, col.names = c("Category", "I want to…", "Run", "Guide"))
```

Prefer the console? `eri_task_map()` prints this same tree interactively, grouped by category, with
every reference function each task touches:

```{r console-version}
eri_task_map()
```
