---
title: "ERI Daily Analyst Workflow"
subtitle: "Data sync, survey health, and approval"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    self-contained: true
knitr:
  opts_chunk:
    echo: true
    message: false
    warning: false
---

<!--
SOP NOTES -- read before running
=================================
1. DO NOT commit this file back to the erifunctions repository.
   It belongs in your analysis project (e.g. under analysis/daily/).

2. Pin the package version you are using:
      renv::snapshot()
   Run this after installing or updating erifunctions.

3. NEVER push directly to main or dev of any shared repo.
   Always use a feature branch and open a pull request.

4. Set all required environment variables in your project .Renviron
   before knitting. To open it:
      usethis::edit_r_environ(scope = "project")
   Required variables:
      ERIFUNCTIONS_TENANT_ID
      ERIFUNCTIONS_APP_ID
      ERIFUNCTIONS_RESOURCE_ENDPOINT
      ERIFUNCTIONS_STORAGE_NAME
      ERIFUNCTIONS_DATA_STORAGE_NAME
      ERI_ANALYST_ID
      ODK_URL
      ODK_USER
      ODK_PASS

5. To get a fresh copy of this template from an updated package:
      file.copy(
        system.file("templates/eri_daily_workflow.qmd", package = "erifunctions"),
        "."
      )
-->

```{r setup}
library(erifunctions)

cat("erifunctions version:", as.character(packageVersion("erifunctions")), "\n")
cat("Run started:", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n")
cat("Analyst:", Sys.getenv("ERI_ANALYST_ID", unset = paste0(Sys.info()[["user"]], " (unverified)")), "\n")
```

```{r connections}
# Azure data container (reads ERIFUNCTIONS_* env vars automatically)
az_con <- get_azure_storage_connection(
  storage_name = Sys.getenv("ERIFUNCTIONS_DATA_STORAGE_NAME", unset = "data")
)

# ODK Central connection (reads ODK_URL / ODK_USER / ODK_PASS env vars)
odk_con <- init_odk_connection()
```

## Survey health check

Check submission recency for all forms you manage. Review the `status`,
`total_submissions`, `submissions_7d`, and `last_submission_at` columns for
anything unexpected before syncing.

```{r survey-status}
# Fetch status for all forms across all projects.
# Scope to a single project:  eri_survey_status(project_id = 7L, con = odk_con)
# Scope to a single form:     eri_survey_status(project_id = 7L, form_id = "MyForm", con = odk_con)
status <- eri_survey_status(con = odk_con)
print(status)
```

## Sync new data

Run `eri_odk_sync()` for each form you manage. Add or remove chunks below as
needed. Each call downloads new submissions and writes them to the Azure data
blob at `{country}/{disease}/research/raw/{form_id}.parquet` (ODK is the
`research` channel's collection format under ADR-0012).

```{r sync-form-1}
# Replace with your actual project_id, form_id, country, and disease values.
# The form must be registered via eri_odk_register() before syncing.
eri_odk_sync(
  project_id = 7L,
  form_id    = "RiverProspection",   # <-- replace
  con        = odk_con,
  data_con   = az_con
)
```

```{r sync-form-2, eval=FALSE}
# Duplicate this chunk for each additional form you manage.
eri_odk_sync(
  project_id = 7L,
  form_id    = "FlyCollection",      # <-- replace
  con        = odk_con,
  data_con   = az_con
)
```

## Review and approve

After syncing, inspect the raw data before promoting it to the processed layer.
`eri_approve()` moves files from `{country}/{disease}/{data_type}/staged/` to
`{country}/{disease}/{data_type}/processed/` and writes an audit log entry.

**Do not approve data you have not reviewed.** Open the staged parquet file
first (e.g. with `eri_read()`), check for obvious issues, then call
`eri_approve()` once you are satisfied.

```{r review-staged, eval=FALSE}
# Preview what is staged before approving.
staged_path <- eri_data_path(
  country     = "uga",         # <-- replace
  disease     = "oncho",       # <-- replace
  data_source = "programmatic",# <-- replace (surveillance, programmatic, odk)
  data_type   = "treatment",   # <-- replace (the measure: case, treatment, tas, ...)
  layer       = "staged"
)
staged_data <- eri_read(staged_path, azcontainer = az_con)
head(staged_data)
```

```{r approve, eval=FALSE}
# Promote staged -> processed.
# period: the data period being approved (e.g. "2026-05" for monthly, "2026-W22" for weekly).
eri_approve(
  country   = "uga",       # <-- replace
  disease   = "oncho",     # <-- replace
  data_type = "cmr",       # <-- replace
  period    = "2026-05"    # <-- replace
)
```

## Session summary

```{r session-summary}
cat("Run completed:", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n")
cat("Analyst:", Sys.getenv("ERI_ANALYST_ID", unset = paste0(Sys.info()[["user"]], " (unverified)")), "\n")
cat("erifunctions version:", as.character(packageVersion("erifunctions")), "\n")
```
