---
title: "ERI Research Project Workflow"
subtitle: "Data pull, analysis, logging, and archiving"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    self-contained: true
knitr:
  opts_chunk:
    echo: true
    message: false
    warning: false
---

<!--
SOP NOTES -- read before running
=================================
1. DO NOT commit this file or data/ back to the erifunctions repository.
   This template belongs in your analysis project repo.

2. Add data/ to your project .gitignore IMMEDIATELY:
      echo "data/" >> .gitignore
   Never commit raw or processed data to git.

3. Pin the package version you are using:
      renv::snapshot()
   Run this after installing or updating erifunctions.

4. Set all required environment variables in your project .Renviron
   before knitting. To open it:
      usethis::edit_r_environ(scope = "project")
   Required variables:
      ERIFUNCTIONS_TENANT_ID
      ERIFUNCTIONS_APP_ID
      ERIFUNCTIONS_RESOURCE_ENDPOINT
      ERIFUNCTIONS_DATA_STORAGE_NAME
      ERI_ANALYST_ID

5. To pull a fresh copy of this template:
      eri_template_pull("eri_research_workflow")

6. One snapshot per major milestone -- before a model run, before submission,
   after peer review. Snapshots are stored in Azure and are not a substitute
   for version control on your analysis code.
-->

```{r setup}
library(erifunctions)

cat("erifunctions version:", as.character(packageVersion("erifunctions")), "\n")
cat("Session started:", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n")
cat("Analyst:", Sys.getenv("ERI_ANALYST_ID", unset = Sys.info()[["user"]]), "\n")
```

## Session setup

On the **first run** of a new project, use `eri_research_init()`. On every
subsequent session, use `eri_research_resume()` -- it re-reads your
`research.yaml` manifest and prints a session summary (last pull, last log
entry, snapshot count) so you can pick up where you left off without
re-typing project context.

```{r init-first-run, eval=FALSE}
# --- FIRST RUN ONLY ---------------------------------------------------------
# Run once to scaffold local dirs (data/, figs/, outputs/) and create the
# Azure project directory. Then switch to eri_research_resume() below.

az_con <- get_azure_storage_connection(
  storage_name = Sys.getenv("ERIFUNCTIONS_DATA_STORAGE_NAME", unset = "data")
)

eri_research_init(
  project_name = "my_project",      # <-- replace: short identifier, no spaces
  country      = "dr",              # <-- replace: country code
  disease      = "malaria",         # <-- replace: disease
  description  = "Brief description of the research question",
  data_con     = az_con
)
```

```{r resume}
# --- EVERY SUBSEQUENT SESSION -----------------------------------------------
# Re-establishes context from research.yaml. Prints last pull, last log,
# snapshot count. Replace with your project root path if not running from it.

az_con <- eri_research_resume()
```

## Pull data

Pull canonical processed data from the three-layer pipeline or any Azure
reference path. Every pull is recorded in `research.yaml` for provenance.

For non-standard external files not yet in Azure (e.g. IRS campaign data,
population grids), upload them first with `eri_artifact_upload()`, then pull
with `eri_artifact_pull()`.

```{r pull-canonical, eval=FALSE}
# Pull processed surveillance data for your country and disease
eri_research_pull(
  country   = "dr",              # <-- replace
  disease   = "malaria",         # <-- replace
  data_type = "surveillance",    # <-- replace: surveillance, cmr, odk
  data_con  = az_con
)
```

```{r pull-spatial, eval=FALSE}
# Pull standard spatial reference files (shapefiles, rasters)
eri_research_pull(
  path     = "spatial/dom_admin_boundaries",  # <-- replace
  dest     = "data/spatial",
  data_con = az_con
)
```

```{r pull-artifact, eval=FALSE}
# Pull a non-standard reference file from the artifact registry
# (upload it first with eri_artifact_upload() if not already registered)
eri_artifact_pull(
  name     = "dr_irs_2024",      # <-- replace: artifact name
  dest     = "data/raw",
  data_con = az_con
)
```

## Analysis

Your analysis code goes here. Keep chunks focused and named. Log key decisions
inline with `eri_research_log()` immediately after they occur -- do not batch
logging at the end of the session.

```{r load-data, eval=FALSE}
# Example: load pulled data
library(arrow)
surveillance <- read_parquet("data/surveillance.parquet")
```

```{r log-data-loaded}
eri_research_log("Data loaded. Surveillance file covers 2020-2024.")
```

```{r analysis, eval=FALSE}
# Your analysis code here.
# Example (ITS model):
#   library(MASS)
#   model <- glm.nb(cases ~ time + intervention + time_after, data = surveillance)
#   summary(model)
```

```{r log-model}
eri_research_log("Ran primary model. Converged. Saving output.")
```

## Archive outputs

Upload key outputs to Azure so they are accessible to the team and versioned
alongside the project manifest. Figures go to `outputs/figs/`, model objects
to `outputs/`.

```{r upload-figure, eval=FALSE}
# Save a figure locally first, then upload
library(ggplot2)
p <- ggplot(surveillance, aes(x = epiweek, y = cases)) + geom_line()
ggsave("figs/its_model.png", p, width = 8, height = 5)

eri_research_upload_figure(
  "figs/its_model.png",
  caption  = "ITS model -- malaria incidence pre/post IRS",  # <-- replace
  data_con = az_con
)
```

```{r upload-output, eval=FALSE}
# Upload R model object for reproducibility
eri_research_upload_output(
  obj      = model,               # <-- replace: any R object
  filename = "its_model.qs2",    # <-- replace
  data_con = az_con
)
```

## Snapshot

Take a snapshot to freeze the full `data/` directory in Azure at a named
milestone. Do this before a major model run, before sharing results, or after
peer review. Snapshots are append-only and recorded in `research.yaml`.

```{r snapshot, eval=FALSE}
eri_research_snapshot(
  label    = "pre-submission",   # <-- replace: short descriptive label
  data_con = az_con
)
```

## Session summary

```{r session-summary}
cat("Session completed:", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n")
cat("Analyst:", Sys.getenv("ERI_ANALYST_ID", unset = Sys.info()[["user"]]), "\n")
cat("erifunctions version:", as.character(packageVersion("erifunctions")), "\n")
```
